---
title: "Synoptic Arctic Survey 2021 - metatranscriptomics and metabarcoding"
author: "george.westmeijer@umu.se"
bibliography: references.bib
csl: diabetologia.csl
date: now
format: html
toc: true
toc_float: true
toc_depth: 3
number_sections: true
theme: sandstone
mainfont: Helvetica
monofont: AndaleMono
fontsize: "8"
highlight: textmate
df_print: paged
code-fold: true
code-summary: "Expand code"
self_contained: true
embed-resources: true
keep_md: false
encoding: "UTF-8"
other-links:
      - text: GitHub repo (private for now)
        href: https://github.com/geweaa/sas-metat
      - text: Nf-core/metatdenovo
        href: https://nf-co.re/metatdenovo/1.0.1/
      - text: Nature guidelines on figures
        href: https://research-figure-guide.nature.com/figures/preparing-figures-our-specifications/
---

# Load libraries

```{r}
#| label: libraries
#| cache: false
#| include: false

library(dplyr)
library(tidyr)
library(tibble)
library(forcats)
library(ggplot2)
library(readr)
library(patchwork)
library(grateful)
```


```{r}
#| label: setup
#| include: false

source("../Protocols/theme.R")
knitr::opts_chunk$set(cache = FALSE)
```


```{r}
#| label: cite-packages
#| cache: false
#| tbl-cap: Versions of R libraries and R itself used during this analysis.

cite_packages(output = "table", pkgs = "Session", out.dir = getwd())
```

# Read input files RNA

The raw data was procesesd using nf-core/metatdenovo (v1.0.1): https://nf-co.re/metatdenovo/1.0.1/ using Prokka as orf caller, GTDB release 220 for taxonomic annotation, and EggNOG mapper for functional annotation.

```{r}
#| label: metatdenovo-gtdb

gtdb <- read_tsv("data/metatdenovo/user_assembly.prokka.gtdb-r220.diamond.taxonomy.tsv.gz", show_col_types = F) %>%
  select(-strain,-taxonomy) %>%
  filter(!is.na(phylum))

# The GTDB metadata 'md' is to join the NCBI with the GTDB taxonomy
pro <- read_tsv("data/metatdenovo/user_assembly.prokka.ncbi-refseq-20250115.diamond.taxonomy-taxdump.tsv.gz", show_col_types = F) %>%
  filter(superkingdom %in% c("Bacteria", "Archaea")) %>%
  select(orf, evalue, taxid) %>%
  filter(!orf %in% gtdb$orf) 

md <- read_tsv("./data/magmap/ar53_metadata_r226.tsv.gz", show_col_types = F)
md <- read_tsv("./data/magmap/bac120_metadata_r226.tsv.gz", show_col_types = F) %>% rbind(md)

pro <- md %>% 
  filter(gtdb_representative) %>%
  select(taxid = ncbi_taxid, gtdb_taxonomy) %>%
  filter(taxid %in% pro$taxid) %>%
  separate_wider_delim(gtdb_taxonomy, delim = ";", names = c("domain", "phylum", "class", "order", "family", "genus", "species")) %>%
  mutate(species = NA) %>%
  mutate(across(-1, \(x) gsub(x, pattern = "[a-z]__", replacement = ""))) %>%
  mutate(genus = gsub("_[A-Z]", "", genus)) %>%
  distinct() %>%
  group_by(taxid) %>% slice_sample(n = 1) %>% ungroup() %>%
  inner_join(pro, ., by = "taxid")

euk <- read_tsv("data/metatdenovo/user_assembly.prokka.ncbi-refseq-20250115.diamond.taxonomy-taxdump.tsv.gz", show_col_types = F) %>%
  filter(superkingdom == "Eukaryota" | superkingdom == "Viruses") %>%
  select(-taxonomy) %>%
  select(orf, taxid, evalue, domain = superkingdom, phylum, class, order, family, genus, species)

# Identify ORFs that are identified as prokaryotes with GTDB but eukaryotes according to NCBI
orfs <- gtdb %>%
  select(orf, gtdb.evalue = evalue) %>%
  inner_join(euk, by = "orf") %>%
  filter(evalue > gtdb.evalue) %>%
  pull(orf)

# Merge into one object: 
tax <- gtdb %>%
  filter(!orf %in% orfs) %>%
  rbind(pro) %>%
  rbind(euk %>% filter(!orf %in% gtdb$orf))

remove(md, orfs, pro)
```


```{r}
#| label: metatdenovo-prokka

tpm <- read_tsv("data/metatdenovo/user_assembly.prokka.counts.tsv.gz", show_col_types = F) %>% 
  mutate(sample = gsub("R", ".", sample)) %>%
  mutate(sample = factor(sample, levels = c(
    "S1.1", "S1.2", "S2.1", "S2.2", "S3.1", "S3.2", "S4.1", "S4.2", "S5.1", "S6.1", 
    "S5.2", "S6.2", "S7.1", "S7.2", "S8.1", "S8.2", "S9.1", "S9.2", "S10.1", "S10.2"
  ))) %>%
  select(-start, -end,- strand, -length, -chr)

meta <- read_tsv("data/metatdenovo/sample_key.tsv", show_col_types = FALSE) %>%
  mutate(station = as.character(station)) %>%
  mutate(sample = gsub("R", ".", sample)) %>%
  mutate(sample = factor(sample, levels = c(
    "S1.1", "S1.2", "S2.1", "S2.2", "S3.1", "S3.2", "S4.1", "S4.2", "S5.1", "S6.1", 
    "S5.2", "S6.2", "S7.1", "S7.2", "S8.1", "S8.2", "S9.1", "S9.2", "S10.1", "S10.2"
  ))) %>%
  mutate(layer = factor(layer, levels = c("10m", "chlMax")))
```

In total, there are `r nrow(meta)` metatranscriptomes from five stations. At every station, samples were taken in duplicates from 10 m depth and from chl. max. On average, each sample has `r round(mean(meta$mreads), digits = 1)` million reads (sd `r sd(meta$mreads) %>% round(digits = 1)`, n = 20). Off all the transcripts, 53.0 % was functionally annotated (Eggnog mapper), 36.5 % was functionally annotated including a KO reference, 51.7 % of the transcripts was annotated on phylum level, 48.0 % on order level, 43.5 % on genus level, and 32.7 % on species level.

63.6 % of the transcript were functionally annotated with Kofam scan. A summary of the bioinformatic pipeline will soon be added to this introduction.

```{r}
#| label: read-eggnogmapper

eggnog <- read_tsv("data/metatdenovo/user_assembly.prokka.emapper.tsv.gz", show_col_types = F) %>%
  mutate(kegg_ko = gsub("ko:", "", kegg_ko)) %>%
  filter(kegg_ko != "-") %>%
  select(orf, evalue, cog = cog_category, ko = kegg_ko, module = kegg_module, name = preferred_name, description, max_annot_lvl)

kofam <- read_tsv("data/metatdenovo/user_assembly.prokka.kofamscan-uniq.tsv.gz", show_col_types = F) %>%
  select(-thrshld, -score)
```

# Read input files DNA

Forward primer 515F (5' GTGYCAGCMGCCGCGGTAA), reverse primer 926R (5' CCGYCAATTYMTTTRAGTTT) from Parada et al. (2015) [-@parada2016]. The data was submitted to ENA under the reference PRJEB63557.

```{r}
#| label: parameters-ampliseq

read_delim("data/ampliseq/params.yaml", delim = " ", col_names = c("parameter", "value"), show_col_types = FALSE) %>%
  knitr::kable()
```


```{r}
#| label: output-ampliseq

seqtab <- read_tsv("data/ampliseq/ASV_table.tsv", show_col_types = FALSE) %>%
  pivot_longer(-1, names_to = "sample", values_to = "count") %>%
  filter(count > 0) %>% mutate(sample = gsub("_S..", "", sample)) %>%
  rename(seqid = ASV_ID) %>%
  group_by(sample) %>% mutate(relab = count / sum(count)) %>% ungroup()
  
asv <- read_tsv("data/ampliseq/ASV_tax_species.sbdi-gtdb.tsv", show_col_types = FALSE) %>%
  rename(seqid = ASV_ID) %>% rename_with(tolower) %>% select(-species_exact, -confidence, -sequence)

dnakey <- read_tsv("data/ampliseq/sample-key.txt", show_col_types = FALSE) %>%
  mutate(station = factor(station, levels = c("5","26","42","53","56"))) %>%
  mutate(layer = factor(layer, levels = c("10m","Chlmax"))) 
```


```{r define colors}
cols.phylum <- c( # Sorted on abundance
  "Pseudomonadota" = "#C7B19C",
  "Thermoplasmatota" = "#972D15",
  "Bacteroidota" = "#FAEFD1",
  "Cyanobacteriota" = "#D69C4E",
  "Thermoproteota" = "#A2A475",
  "Marinisomatota" = "#00A08A",
  "Chloroflexota" = "#D8B70A",
  "SAR324" = "#1B5331",
  "Other" = "#D5D5D3",
  "Unidentified" = "#8D8680"
)

cols.order <- c( # Sorted on abundance
  "Poseidoniales" = "#D69C4E",
  "Pseudomonadales" = "#C7B19C",
  "Pelagibacterales" = "#972D15",
  "Flavobacteriales" = "#D8B70A",
  "PS1" = "#FAEFD1",
  "Nitrososphaerales" = "#A2A475",
  "Enterobacterales" = "#00A08A",
  "SAR86" = "#1B5331",
  "Rhodobacterales" = "#9986A5",
  "Marinisomatales" = "#79402E",
  "Puniceispirillales_A" = "#046C9A",
  "Other" = "#D5D5D3"
)

cols.genus <- c( # Sorted on abundance
  "MGIIb-O3" = "#C7B19C",
  "Pseudothioglobus" = "#A2A475",
  "Pelagibacter" = "#972D15",
  "UBA9145" = "#D69C4E",
  "HTCC2207" = "#FAEFD1",
  "Lutibacter" = "#00A08A",
  "Photorhabdus" = "#D8B70A",
  "Nitrosopumilus" = "#1B5331",
  "ASP10-02a" = "#046C9A",
  "Arctic96AD-7" = "#D5D5D3"
)

cols.layer <- c(
  "10m" = "#78B7C5",
  "chlMax" = "#046C9A"
)

cols.station <- c(
   "5" = "#046C9A",
  "26" = "#446455",
  "42" = "#FDD262",
  "53" = "#D3DDDC",
  "56" = "#C7B19C"
)
```

# Data exploration

## Rarefaction

```{r}
#| label: rarefaction-amplicons
# The lowest number of reads in a sample is 4.1 million

rare <- tpm %>%
  select(-tpm) %>%
  pivot_wider(names_from = "orf", values_from = "count", values_fill = 0) %>%
  column_to_rownames("sample") %>%
  vegan::rarecurve(sample = 4.1e6, step = 1e5, tidy = TRUE) %>%
  tibble() %>%
  rename_with(tolower)
```

```{r}
#| fig-cap: Rarefaction curves to determine sufficient sequencing depth of the RNA extract
#| label: fig-rarefaction
#| out-width: 100%

rare %>%
  rename(samplesize = sample, sample = site, n.orfs = species) %>%
  inner_join(meta, by = "sample") %>%
  ggplot(aes(
    x = samplesize, 
    y = n.orfs,
    colour = layer
    )) + 
  geom_line() + scale_color_manual(
    values = c("10m" = "#899DA4", "chlMax" = "#046C9A"),
    labels = c("10m" = "10 m", "chlMax" = "Chl. max")
    ) +
  labs(x = "No. of sequence reads", y = "Rarefied No. of ORFs") +
  #scale_x_continuous(labels = c("0", "50.000", "100.000", "150.000")) +
  theme_tidy(fontsize = 6) +
  theme(
    legend.title = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.85, 0.15)
  )
```

## Percentage of annotated transcripts

```{r}
#| label: Percentage of transcripts with a taxonomic annotation
#| tbl-cap: Percentage of transcripts with a taxonomic annotation from both GTDB and NCBI

i <- c("domain","phylum","class","order","family","genus")

for (level in i) {
  ## The (!!as.name(level)) converts the variable level to e.g. 'genus'
  d <- tpm %>% left_join(tax, by = "orf") %>% filter(!is.na((!!as.name(level))))
  format(nrow(d) / nrow(tpm) * 100, digits = 3) %>% paste(stringr::str_to_title(level), ": ", ., " %", sep = "") %>% print()
}
```

```{r}
#| label: Percentage of transcripts with a taxonomic and functional annotation (KO)
#| tbl-cap: Percentage of transcripts with a taxonomic and a functional annotation

i <- c("domain","phylum","class","order","family","genus")

for (level in i) {
  ## The (!!as.name(level)) converts the variable level to e.g. 'genus'
  d <- tpm %>% left_join(tax, by = "orf") %>% inner_join(kofam, by = "orf") %>% filter(!is.na((!!as.name(level))))
  format(nrow(d) / nrow(tpm) * 100, digits = 3) %>% paste(stringr::str_to_title(level), ": ", ., " %", sep = "") %>% print()
}
```


## Community composition on domain level

```{r}
#| label: tbl-domain

tpm %>%
  left_join(tax, by = "orf") %>%
  replace_na(list("domain" = "Unknown")) %>%
  group_by(domain) %>% summarise(tpm = sum(tpm) / 20, .groups = "drop") %>%
  arrange(desc(tpm))
```



@fig-domain depicts the ratio between bacteria and archaea for both depth layers. Note how the abundance of archaea is slightly higher in most of the deeper samples (chlorophyll max depth).

```{r}
#| fig-cap: Community overview on domain level, removing transcripts without a taxonomic annotation.
#| label: fig-domain
#| out-width: 100%
#| fig-width: 4.75

tpm %>%
  inner_join(tax, by = "orf") %>%
  inner_join(meta, by = "sample") %>%
  group_by(sample, domain) %>% summarise(tpm = sum(tpm), .groups = "drop") %>%
  mutate(domain = factor(domain, levels = c("Viruses", "Eukaryota", "Archaea", "Bacteria"))) %>%
  ggplot(aes(
    x = sample,
    y = tpm,
    fill = domain
  )) +
  geom_col() + theme_tidy() +
  scale_y_continuous(breaks = c(0, 2e5, 4e5, 6e5), labels = c(0, 0.2, 0.4, 0.6)) +
  labs(y = expression('Relative abundance (tpm '%*% '10'^'6'*')')) + 
  scale_fill_manual(
    values = c("Bacteria" = "#C7B19C", "Archaea" = "#A2A475", "Viruses" = "#D8B70A", "Eukaryota" = "#FAEFD1")
    ) +
  theme(
    aspect.ratio = 0.5,
    legend.title = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  ) +
  annotate("segment", x =  0.55, xend =  2.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x =  2.55, xend =  4.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x =  4.55, xend =  6.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x =  6.55, xend =  8.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x =  8.55, xend = 10.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 10.55, xend = 12.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x = 12.55, xend = 14.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 14.55, xend = 16.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x = 16.55, xend = 18.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 18.55, xend = 20.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("label", x = 2.5, y = -30000, label = "5", label.size = 0, size = 6/.pt) +
  annotate("label", x = 6.5, y = -30000, label = "26", label.size = 0, size = 6/.pt) +
  annotate("label", x =10.5, y = -30000, label = "42", label.size = 0, size = 6/.pt) +
  annotate("label", x =14.5, y = -30000, label = "53", label.size = 0, size = 6/.pt) +
  annotate("label", x =18.5, y = -30000, label = "56", label.size = 0, size = 6/.pt)
```

# Community composition

## Community composition - phylum

```{r most abundant phyla for plotting}
#| tbl-cap: The eight most abundant phyla of all samples, based on summed TPM. Most are bacterial phyla, except for the archaeal phyla Thermoproteota and Thermoplasmatota. 

t <- tpm %>%
  inner_join(tax, by = "orf") %>% filter(domain %in% c("Bacteria", "Archaea")) %>%
  group_by(phylum, sample) %>%
  # Sum the abundance of each phylum within a sample
  summarise(tpm = sum(tpm), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean.tpm = sum(tpm), .groups = 'drop') %>%
  filter(!is.na(phylum)) %>%
  top_n(8, mean.tpm)

t %>% arrange(desc(mean.tpm))
```


```{r prepare the taxonomy, include=FALSE}
cols.phylum <- c( # Sorted on abundance
  "Pseudomonadota" = "#C7B19C",
  "Thermoplasmatota" = "#972D15",
  "Bacteroidota" = "#FAEFD1",
  "Cyanobacteriota" = "#D69C4E",
  "Thermoproteota" = "#A2A475",
  "Marinisomatota" = "#00A08A",
  "Chloroflexota" = "#D8B70A",
  "SAR324" = "#1B5331",
  "Other" = "#D5D5D3",
  "Eukaryotes" = "#046C9A"
  #"Unidentified" = "#8D8680"
)
```


@fig-phylum explores the data by plotting the most abundant phyla over all samples, regardless if the RNA transcript was functionally annotated or not. Note that the taxonomy is outdated and needs to be updated (e.g. Pseudomonadota instead of Proteobacteria). 87.7 % of the annotated ORFs are affiliated with bacteria, while 13.1 % is affiliated with Archaea.

```{r}
#| label: fig-phylum
#| fig-cap: Community composition of all RNA transcripts on phylum level. The phyla are sorted according to relative abundance while grouping low-abundant phyla as 'Other' and transcripts without any taxonomic annotation as 'Unidentified'
#| out-width: 100%

p.phylum <- tpm %>%
  left_join(tax, by = "orf") %>%
  mutate(phylum = case_when(
    phylum %in% t$phylum ~ phylum,
    domain == "Eukaryota" ~ "Eukaryotes",
    is.na(phylum) ~ "Unidentified",
    !phylum %in% t$phylum ~ "Other"
  )) %>%
  filter(phylum != "Unidentified") %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(sample, phylum) %>% 
  summarise(tpm = sum(tpm), .groups = 'drop') %>%
  # Call the plot
  ggplot(aes(x = sample, 
             y = tpm, 
             fill = fct_relevel(phylum, names(cols.phylum) %>% rev())
             )
         ) +
  labs(x = '', y = expression('Relative abundance (tpm '%*% '10'^'6'*')'), fill = "") +
  geom_col() +
  scale_y_continuous(breaks = c(0, 2e5, 4e5, 6e5, 8e5, 1e6),
                labels = c("0.0", "0.2", "0.4", "0.6", "0.8", "1.0")
                ) +
  scale_fill_manual(values = cols.phylum) +
  theme_tidy(fontsize = 5) + 
  theme(
    aspect.ratio = 0.5,
    axis.ticks.x = element_blank(),
    legend.key.height = unit(3.5, "mm"),
    legend.margin = margin(),
    panel.border = element_rect(fill = "transparent", linewidth = 0.5),
    axis.text.x = element_blank()
    ) +
  annotate("segment", x =  0.55, xend =  2.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x =  2.55, xend =  4.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x =  4.55, xend =  6.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x =  6.55, xend =  8.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x =  8.55, xend = 10.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 10.55, xend = 12.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x = 12.55, xend = 14.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 14.55, xend = 16.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x = 16.55, xend = 18.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 18.55, xend = 20.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("label", x = 2.5, y = -30000, label = "5", label.size = 0, size = 5/.pt) +
  annotate("label", x = 6.5, y = -30000, label = "26", label.size = 0, size = 5/.pt) +
  annotate("label", x =10.5, y = -30000, label = "42", label.size = 0, size = 5/.pt) +
  annotate("label", x =14.5, y = -30000, label = "53", label.size = 0, size = 5/.pt) +
  annotate("label", x =18.5, y = -30000, label = "56", label.size = 0, size = 5/.pt)

p.phylum
```

```{r}
ggsave(filename = "figures/metatdenovo-phylum.pdf", width = 180, height = 120, units = "mm")
```


## Community composition - order

```{r most abundant orders for plotting}
#| tbl-cap: The eight most abundant orders, after summing the TPM of all the samples.

t <- tpm %>%
  inner_join(tax, by = "orf") %>% filter(domain %in% c("Bacteria", "Archaea")) %>%
  group_by(order, sample) %>%
  # Sum the abundance of each phylum within a sample
  summarise(tpm = sum(tpm), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean.tpm = sum(tpm), .groups = 'drop') %>%
  filter(!is.na(order)) %>%
  top_n(9, mean.tpm)

t %>% arrange(desc(mean.tpm))
```


```{r prepare the taxonomy - order, include=FALSE}
cols.order <- c( # Sorted on abundance
  "Pseudomonadales" = "#C7B19C",
  "Poseidoniales" = "#D69C4E",
  "Pelagibacterales" = "#972D15",
  "PS1" = "#FAEFD1",
  "Flavobacteriales" = "#D8B70A",
  "Nitrososphaerales" = "#A2A475",
  "Enterobacterales" = "#00A08A",
  "SAR86" = "#1B5331",
  "Rhodobacterales" = "#9986A5",
  "Other" = "#D5D5D3",
  "Eukaryotes" = "#046C9A"
)
```

@fig-order explores the data by plotting the most abundant orders over all samples, regardless if the RNA transcript was functionally annotated or not. Note that transcripts without a taxonomic annotation (the unidentified fraction in @fig-phylum) are not included in this figure. It clearly shows how the Nitrosphaerales are more abundant in the deeper layer.

```{r}
#| label: fig-order
#| fig-cap: Community composition of all RNA transcripts on order level. The orders are sorted according to tpm while grouping low-abundant orders as 'Other'. Transcripts without taxonomic annotation are not included. For each station, the two bars left of the label were sampled at 10 m depth, while the two bars on the right were sampled at chlorophyll max (between 20 and 45 m depth)
#| out-width: 100%

p.order <- tpm %>%
  left_join(tax, by = "orf") %>%
  mutate(order = case_when(
    order %in% t$order ~ order,
    domain == "Eukaryota" ~ "Eukaryotes",
    is.na(order) ~ "Unidentified",
    !order %in% t$order ~ "Other"
  )) %>%
  filter(order != "Unidentified") %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(sample, order) %>% 
  summarise(tpm = sum(tpm), .groups = 'drop') %>%
  # Call the plot
  ggplot(aes(x = sample, 
             y = tpm, 
             fill = fct_relevel(order, rev(names(cols.order)))
             )
         ) +
  labs(x = '', y = expression('Relative abundance (tpm '%*% '10'^'6'*')'), fill = "Order") +
  geom_col() +
  scale_y_continuous(breaks = c(0, 2e5, 4e5, 6e5, 8e5, 1e6),
                labels = c("0.0", "0.2", "0.4", "0.6", "0.8", "1.0")
                ) +
  scale_fill_manual(values = cols.order) +
  theme_tidy(legend = "right", fontsize = 5) +
  #guides(fill = guide_legend(nrow = 3, reverse = TRUE)) +
  theme(
    aspect.ratio = 0.75,
    axis.ticks.x = element_blank(),
    legend.key.width = unit(3.5, "mm"),
    legend.key.height = unit(4, "mm"),
    legend.margin = margin(t = -10),
    legend.text = element_text(margin = margin(l = 2)),
    axis.text.x = element_blank()
    ) +
  annotate("segment", x =  0.55, xend =  2.45, y = -25000, yend = -25000, color = "#78B7C5") +
  annotate("segment", x =  2.55, xend =  4.45, y = -25000, yend = -25000, color = "#046C9A") +
  annotate("segment", x =  4.55, xend =  6.45, y = -25000, yend = -25000, color = "#78B7C5") +
  annotate("segment", x =  6.55, xend =  8.45, y = -25000, yend = -25000, color = "#046C9A") +
  annotate("segment", x =  8.55, xend = 10.45, y = -25000, yend = -25000, color = "#78B7C5") +
  annotate("segment", x = 10.55, xend = 12.45, y = -25000, yend = -25000, color = "#046C9A") +
  annotate("segment", x = 12.55, xend = 14.45, y = -25000, yend = -25000, color = "#78B7C5") +
  annotate("segment", x = 14.55, xend = 16.45, y = -25000, yend = -25000, color = "#046C9A") +
  annotate("segment", x = 16.55, xend = 18.45, y = -25000, yend = -25000, color = "#78B7C5") +
  annotate("segment", x = 18.55, xend = 20.45, y = -25000, yend = -25000, color = "#046C9A") +
  annotate("label", x = 2.5, y = -25000, label = "Station 5", label.size = 0, size = 5/.pt) +
  annotate("label", x = 6.5, y = -25000, label = "26", label.size = 0, size = 5/.pt) +
  annotate("label", x =10.5, y = -25000, label = "42", label.size = 0, size = 5/.pt) +
  annotate("label", x =14.5, y = -25000, label = "53", label.size = 0, size = 5/.pt) +
  annotate("label", x =18.5, y = -25000, label = "56", label.size = 0, size = 5/.pt)

p.order
```

## Community composition - genus

```{r}
#| label: table-genus
#| tbl-cap: Twenty most abundant genera over all samples, based on tpm. The order and phylum of the respective genera are also displayed.  

t <- tpm %>%
  inner_join(tax, by = "orf") %>% filter(domain %in% c("Bacteria", "Archaea")) %>%
  group_by(genus, sample) %>%
  # Sum the abundance of each phylum within a sample
  summarise(tpm = sum(tpm), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean.tpm = sum(tpm), .groups = 'drop') %>%
  filter(!is.na(genus)) %>%
  top_n(10, mean.tpm)

t %>% left_join(tax, by = "genus") %>% 
  select(genus, order, phylum, mean.tpm) %>% 
  distinct() %>%
  arrange(desc(mean.tpm)) %>%
  rename(Genus = genus, Order = order, Phylum = phylum) %>%
  knitr::kable()
```

```{r, include=FALSE}
cols.genus <- c( # Sorted on abundance
  "MGIIb-O3" = "#C7B19C",
  "Pseudothioglobus" = "#A2A475",
  "Pelagibacter" = "#972D15",
  "Nitrosopumilus" = "#1B5331",
  "HTCC2207" = "#FAEFD1",
  "UBA9145" = "#D69C4E",
  "Lutibacter" = "#00A08A",
  "Photorhabdus" = "#D8B70A",
  "ASP10-02a" = "#046C9A",
  "Arctic96AD-7" = "#9986A5",
  "Other" = "#D5D5D3"
)
```

@fig-genus shows the most abundant genera. It reveals how some genera (e.g. Pelagibacter and Ketobacter) are more abundant (and active) in the 10 m depth, while for example Nitrosopumilus is more abundant in the deeper sites. Note from the metadata that the depth of chlorophyll max strongly varies between the stations.

```{r}
#| label: fig-genus
#| fig-cap: Community composition of all RNA transcripts on genus level. Only the top ten most abundant genera are shown, accounting for ~ 20 % of the tpm on average.
#| out-width: 100%

p.genus <- tpm %>%
  left_join(tax, by = "orf") %>%
  mutate(genus = case_when(
    genus %in% t$genus ~ genus,
    is.na(genus) ~ "Unidentified",
    !genus %in% t$genus ~ "Other"
  )) %>%
  filter(genus != "Unidentified") %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(sample, genus) %>% 
  summarise(tpm = sum(tpm), .groups = 'drop') %>%
  # Call the plot
  ggplot(aes(x = sample, 
             y = tpm, 
             fill = fct_relevel(genus, rev(names(cols.genus)))
             )
         ) +
  labs(x = '', y = expression('Relative abundance (tpm '%*% '10'^'6'*')'), fill = "Genus") +
  geom_col() +
  scale_y_continuous(breaks = c(0, 2e5, 4e5, 6e5),
                labels = c("0.0", "0.2", "0.4", "0.6")
                ) +
  scale_fill_manual(values = cols.genus) +
  theme_tidy(legend = "right", fontsize = 5) +
  theme(
    aspect.ratio = 0.5,
    axis.ticks.x = element_blank(),
    legend.key.height = unit(3.5, "mm"),
    axis.text.x = element_blank(),
    legend.title = element_blank(),
    legend.margin = margin()
    ) +
  annotate("segment", x =  0.55, xend =  2.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x =  2.55, xend =  4.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x =  4.55, xend =  6.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x =  6.55, xend =  8.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x =  8.55, xend = 10.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 10.55, xend = 12.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x = 12.55, xend = 14.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 14.55, xend = 16.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("segment", x = 16.55, xend = 18.45, y = -30000, yend = -30000, color = "#78B7C5") +
  annotate("segment", x = 18.55, xend = 20.45, y = -30000, yend = -30000, color = "#046C9A") +
  annotate("label", x = 2.5, y = -30000, label = "5", label.size = 0, size = 5/.pt) +
  annotate("label", x = 6.5, y = -30000, label = "26", label.size = 0, size = 5/.pt) +
  annotate("label", x =10.5, y = -30000, label = "42", label.size = 0, size = 5/.pt) +
  annotate("label", x =14.5, y = -30000, label = "53", label.size = 0, size = 5/.pt) +
  annotate("label", x =18.5, y = -30000, label = "56", label.size = 0, size = 5/.pt)

p.genus
```

```{r}
p.genus / p.phylum + plot_annotation(tag_levels = "a") & 
  theme(
    plot.tag.location = "panel",
    plot.tag.position = c(0.03, 0.95)
  )
```


```{r}
ggsave(filename = "figures/community-genus-phylum.pdf", width = 180, height = 160, units = "mm")
```


## Redundancy analysis (RDA)

```{r}
hydro <- readxl::read_excel("data/primarydata.xlsx", sheet = 1) %>%
  mutate(station = as.character(station)) %>%
  inner_join(meta, by = c("station", "layer", "replicate")) %>%
  rename_with(tolower) %>%
  mutate(sample = factor(sample, levels = c(
    "S1.1", "S1.2", "S2.1", "S2.2", "S3.1", "S3.2", "S4.1", "S4.2", "S5.1", "S6.1", 
    "S5.2", "S6.2", "S7.1", "S7.2", "S8.1", "S8.2", "S9.1", "S9.2", "S10.1", "S10.2"
  ))) %>% arrange(sample)
```


```{r}
hellinger <- tpm %>%
  inner_join(kofam, by = "orf") %>% group_by(sample, ko) %>% summarise(tpm = sum(tpm), .groups = "drop") %>%
  # Standard Vegan transformation: Turn table with with samples as *rows*
  select(sample, ko, tpm) %>%
  pivot_wider(names_from = "ko", values_from = "tpm", values_fill = 0) %>%
  # Turn into a numeric matrix
  column_to_rownames('sample') %>% vegan::decostand(method = "hellinger") %>%
  rownames_to_column("sample") %>%
  pivot_longer(-1, names_to = "ko", values_to = "hellinger")

m <- hellinger %>% # Create a matrix
  spread(ko, hellinger) %>%
  mutate(sample = factor(sample, levels = c(
    "S1.1", "S1.2", "S2.1", "S2.2", "S3.1", "S3.2", "S4.1", "S4.2", "S5.1", "S6.1", 
    "S5.2", "S6.2", "S7.1", "S7.2", "S8.1", "S8.2", "S9.1", "S9.2", "S10.1", "S10.2"
  ))) %>% arrange(sample) %>%
  column_to_rownames("sample")
```


```{r rda}
set.seed(999)

vegan::rda(
  m ~ depth + doc + tdp + tdn + chlach, 
  correlation = T,
  data = hydro %>%
    column_to_rownames('sample') 
) -> rda

# This one will contain the proportions explained which we get by dividing the
# eigenvalue of each component with the sum of eigenvalues for all components.
p <- c(rda$CCA$eig, rda$CA$eig)/sum(c(rda$CCA$eig, rda$CA$eig))
# This one is collecting the coordinates for arrows that will depict how the 
# factors in our model point in the coordinate
a <- rda$CCA$biplot %>% data.frame() %>% tibble::rownames_to_column('factor')

lab <- tibble(factor = a$factor,
            label = c("Depth", "", "TDP & DOC", "TDN", "Chl.")
) %>% inner_join(a, by = "factor")

summary(rda)
vegan::RsquareAdj(rda)
```

```{r}
p.rda <- rda$CCA$wa %>% data.frame() %>% tibble::rownames_to_column('sample') %>%
  inner_join(hydro, by = 'sample') %>%
  ggplot(aes(x = RDA1, y = RDA2)) +
  geom_vline(xintercept = 0, colour = "grey", linetype = "dotted") +
  geom_hline(yintercept = 0, colour = "grey", linetype = "dotted") +
  geom_point(aes(colour = layer, fill = layer), size = 5, shape = 21, stroke = 0.75) +
  xlab(sprintf("RDA1 (%2.1f%% explained)", p[[1]] * 100)) +
  ylab(sprintf("RDA2 (%2.1f%% explained)", p[[2]] * 100)) +
  geom_text(aes(label = station), size = 5/.pt, color = "black") +
  geom_segment(
    data = a, mapping = aes(xend = RDA1/4, yend = RDA2/4), linewidth = 0.3,
    x = 0, y = 0, arrow = arrow(length = unit(1.5, "mm"), type = "closed") 
    ) +
  geom_label(
    data = lab, mapping = aes(x = RDA1/8, y = RDA2/8, label = label),
    size = 5/.pt, parse = F, label.size = 0
    ) + 
  # Add R-squared
  annotate("text", x = Inf, y = -Inf, 
           label = "paste(R ^ 2-adj, \" = 0.56\")", 
           size = 5/.pt, vjust = -0.5, hjust = 1.05, parse = TRUE) +
  scale_color_manual(values = cols.layer, labels = c("10 m", "Chl. max"), name = "") +  
  scale_fill_manual(values = alpha(cols.layer, alpha = 0.2), guide = "none") +      
  theme_tidy(fontsize = 5) +
  theme(
    legend.background = element_rect(fill = NA),
    legend.text = element_text(margin = margin(r = 12, unit = "pt")),
    legend.position = "inside",
    legend.position.inside  = c(0.89,0.93),
    aspect.ratio = 1.0
    )

p.rda
```

```{r}
#| label: fig-nmdstax

(p.order + p.rda) + plot_annotation(tag_levels = "a") & theme(
  plot.margin = margin(r = 4),
  plot.tag.location = "panel",
  plot.tag.position = c(0.03, 0.97),
  )
```

```{r}
#| label: export-rda

ggsave(filename = "figures/rda.pdf", width = 180, height = 80, units = "mm")
```

## Broad overview community functions using COG categories

```{r}
#| label: read-cogs

cogs <- read_tsv("data/cogs.tsv", show_col_types = FALSE) %>%
  rename(cog = cog_category)

i <- eggnog %>% inner_join(cogs, by = "cog") %>%
  inner_join(tpm, by = "orf") %>%
  inner_join(meta, by = "sample") %>% filter(station != "5") %>%
  group_by(cat) %>% summarise(tpm = sum(tpm), .groups = "drop") %>% arrange(desc(tpm)) %>% pull(cat)
```


```{r, fig.width=18/2.53}
#| fig-cap: The abundance (in tpm) of COG categories summed for each depth layer.
#| label: fig-cog

eggnog %>% inner_join(cogs, by = "cog") %>%
  inner_join(tpm, by = "orf") %>%
  inner_join(meta, by = "sample") %>% filter(station != "5") %>%
  group_by(layer, cat) %>% summarise(tpm = sum(tpm), .groups = "drop") %>%
  mutate(layer = factor(layer, levels = c("chlMax","10m"))) %>%
  # Plot
  ggplot(aes(
    x = tpm,
    y = fct_relevel(cat, i),
    group = layer,
    fill = layer
  )) +
  geom_col(width = 0.7, position = position_dodge(width = 0.8)) + 
  scale_fill_manual(values = c("chlMax" = "#046C9A", "10m" = "#78B7C5"), 
                    labels = c("10m" = "10 m", "chlMax" = "Chl. max")
                    ) +
  theme_tidy() + 
  scale_x_continuous(breaks = c(0, 2e5, 4e5, 6e5), labels = c(
    "0", 
    expression("2" %*% "10"^5), expression("4" %*% "10"^5), expression("6" %*% "10"^5)
    )) + xlab("Relative abundance (tpm)") + guides(fill = guide_legend(reverse = T)) +
  theme(
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.80,0.85),
    legend.background = element_blank(),
    aspect.ratio = 1.5
  )
```

# Functional annotation - most abundant KOs

```{r}
#| tbl-cap: The most abundant KEGG orthologs (KO), their description (from Eggnog), and the summed tpm of each KO.
#| label: abundant-kos
 
kofam %>%
  inner_join(tpm, by = "orf") %>%
  group_by(ko, ko_definition) %>% summarise(tpm = sum(tpm), .groups = "drop") %>%
  slice_max(tpm, n = 100) %>%
  knitr::kable(digits = 0, format = "simple")
```

The gene list is from Royo-Llonch et al. [-@royo2021compendium]. It contains the KEGG ortholog (KO) and the pathway the gene is involved in. KOs of interest in the data that are not represented in the gene list are manually categorized. Shows the RNA transcripts for the carbon fixation pathways for each depth layer. To do so, one KO catalyzing the rate-limiting reaction was selected for each pathway. The Wood-Ljungdahl pathway was also included in this analysis but the representing KO was not detected, or the detected KO did not have a taxonomic annotation.

```{r}
#| label: palette-kos

cols.order <- c( # Sorted on abundance
  "Nitrososphaerales" = "#A2A475",
  "Cyanobacteriales" = "#D69C4E",
  "Flavobacteriales" = "#D8B70A",
  "Burkholderiales" = "#78B7C5",
  "Pseudomonadales" = "#C7B19C",
  "Chitinophagales" = "#00A08A",
  "Rhodobacterales" = "#046C9A",
  "Campylobacterales" = "#1B5331",
  "Pelagibacterales" = "#972D15",
  "PS1" = "#FAEFD1",
  "Other" = "#D5D5D3",
  "Unidentified" = "#8D8680"
) %>% rev()
```


```{r}
#| label: read-phosphorus

r <- read_delim("data/p.tsv", col_types = 'ccc') %>% 
  filter(Group == "Remodelling") %>%
  pull(ko)
```




```{r}
#| label: filter-kos

kos <- kofam %>% 
  filter(evalue < 1e-3) %>%
  separate_longer_delim(ko, delim = ",") %>%  
  mutate(pathway = case_when(
    # C fixation
    ko == "K01601" ~ "Calvin cycle: C fixation", # ribulose-bisphosphate carboxylase large chain - EC:4.1.1.39 - Calvin cycle
    ko == "K01602" ~ "Calvin cycle: C fixation", # ribulose-bisphosphate carboxylase small chain - EC:4.1.1.39 - Calvin cycle
    ko == "K01965" ~ "3-hydroxypropionate bi-cycle: C fixation", # biotin dependent acetyl CoA propionyl-CoA carboxylase
    ko == "K01966" ~ "3-hydroxypropionate bi-cycle: C fixation", # biotin dependent acetyl CoA propionyl-CoA carboxylase
    ko == "K00175" ~ "Reverse TCA cycle: C fixation", # 2-oxoglutarate/2-oxoacid ferredoxin oxidoreductase subunit beta - EC:1.2.7.3 1.2.7.11 - Reverse TCA cycle
    ko == "K15230" ~ "Reverse TCA cycle: C fixation", 
    ko == "K15231" ~ "Reverse TCA cycle: C fixation", 
    ko == "K15019" ~ "HP/HB cycle: C fixation", # 3-hydroxypropionyl-coenzyme A dehydratase - EC:4.2.1.116 - HP/HB cycle
    ko == "K14534" ~ "DC/HB cycle: C fixation", # 4-hydroxybutyryl-CoA dehydratase / vinylacetyl-CoA-Delta-isomerase - EC:4.2.1.120 5.3.3.3 - DC/HB cycle
    ko == "K00198" ~ "Wood-Ljungdahl pathway: C fixation",  # anaerobic carbon-monoxide dehydrogenase catalytic subunit - EC:1.2.7.4 - Wood-Ljungdahl
    # N metabolism
    ko %in% c("K00368","K15864","K04561","K02305","K00376","K15877") ~ "Denitrification: N metabolism",
    ko %in% c("K10535","PF01774","PF00449","K00370") ~ "Nitrification: N metabolism",
    ko %in% c("K17877","K00366","K10534","K00367","K00360","K00372") ~ "Assimilatory nitrate reduction: N metabolism",
    ko %in% c("K00362","K00363","K03385","K15876") ~ "Dissimilatory nitrite reduction: N metabolism",
    ko %in% c("K10944","K10945","K10946") ~ "Ammonia monooxygenase (AmoABC): N metabolism",
    ko == "K03320" ~ "Ammonium transporter (amt): N metabolism",
    ko %in% c("K02586","K02588","K02591") ~ "Nitrogen fixation (nifDHK): N metabolism",
    ko %in% c("K01915", "K00264", "K00265", "K00266") ~ "Ammonia assimilation: N metabolism",
    # Central C
    ko == "K01637" ~ "Glyoxylate cycle (ICL): Central C metabolism",
    ko == "K13937" ~ "Pentose phosphate pathway (G6PD): Central C metabolism",
    ko == "K00850" ~ "Glycolysis (PFK): Central C metabolism",
    ko == "K01690" ~ "ED pathway (edd): Central C metabolism",
    ko == "K00031" ~ "Citric acid cycle (icd): Central C metabolism",
    ko == "K18981" ~ "Uronate pathway (udh): Central C metabolism",
    ko == "K00948" ~ "PRPP biosynthesis (RPPK): Central C metabolism",    
    # Photosystems
    ko == "K02703" ~ "Photosystem II (psbA): Photosystems and ATP",
    ko == "K02712" ~ "Photosystem II (psbK): Photosystems and ATP",
    ko == "K02707" ~ "Photosystem II (psbE): Photosystems and ATP",
    ko == "K02697" ~ "Photosystem I (psaJ): Photosystems and ATP",
    ko == "K02274" ~ "Cytochrome c oxidase (coxA): Photosystems and ATP",
    ko == "K02256" ~ "Cytochrome c oxidase (Cox1): Photosystems and ATP",
    # S metabolism
    ko %in% c("K00955", "K00956", "K00957", "K00958") ~ "Sulfate adenylyltransferase (sat): S metabolism",
    ko %in% c("K00394", "K00395") ~ "Adenylylsulfate reductase (aprAB): S metabolism", # Dissimilatory sulfate red.
    ko %in% c("K16950", "K16951", "K00385") ~ "Anaerobic sulfite reductase: S metabolism", # Dissimilatory sulfate red.
    ko %in% c("K21307", "K21308", "K21309") ~ "Sulfite dehydrogenase (quinone): S metabolism", 
    ko %in% c("K17222","K17223","K17224","K17225","K17226","K17227") ~ "Sulfur oxidation (soxAXBCYZ): S metabolism",
    # Metabolism
    ko %in% c("K02043","K22390","K01077","K03430","K05306","K05780","K06162","K06167") ~ "P degradation: P metabolism",
    ko %in% c("K02039","K06217","K07175","K07636","K07657","K07658","K07768","K07776") ~ "P regulation: P metabolism",
    ko %in% c("K08176","K03306","K02036","K02037","K02038","K02040","K02041","K02042","K02044","K14430","K16322") ~ "P transport: P metabolism",
    ko %in% c("K00937","K22468","K01524","K23753") ~ "P stock/destock: P metabolism",
    ko %in% r ~ "P remodelling: P metabolism",
    .default = NA
  )) %>% filter(!is.na(pathway)) %>%
  separate_wider_delim(pathway, delim = ": ", names = c("pathway", "metabolism")) %>%
  left_join(tax, by = "orf") %>%
  replace_na(list("domain" = "Bacteria", "order" = "Unidentified", "phylum" = "Unidentified")) %>%
  filter(domain %in% c("Bacteria", "Archaea")) %>%
  inner_join(tpm, by = "orf", relationship = "many-to-many") %>%
  inner_join(meta, by = "sample") %>%
  mutate(order = if_else(order %in% names(cols.order), order, "Other")) %>%
  mutate(order = factor(order, levels = names(cols.order))) %>%
  mutate(pathway = if_else(
     layer == "10m", true = pathway, paste(pathway, layer)
   )) %>%
  group_by(pathway, metabolism, layer, order) %>% summarise(tpm = sum(tpm), .groups = "drop")
```



```{r}
#| label: fig-fixation

p1 <- kos %>% 
  filter(metabolism == "C fixation") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(pathway),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = kos %>% 
                     filter(metabolism == "C fixation") %>% 
                     select(pathway) %>% distinct() %>% 
                     mutate(pathway = if_else(grepl("chlMax", pathway), "", pathway)) %>%
                     pull(pathway) %>% rev()
                     ) +
  labs(x = "", y = "") + 
  theme_tidy(fontsize = 5) 

p1
```


```{r}
#| label: fig-central-c

p2 <- kos %>% 
  filter(metabolism == "Central C metabolism") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(pathway),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = kos %>% 
                     filter(metabolism == "Central C metabolism") %>% 
                     select(pathway) %>% distinct() %>% 
                     mutate(pathway = if_else(grepl("chlMax", pathway), "", pathway)) %>%
                     pull(pathway) %>% rev()
                     ) +
  labs(x = "", y = "") + 
  theme_tidy(fontsize = 5)

p2
```

The genes listed below are typically catalyze the rate limiting step of the pathway. Like all other analysis, filtering is done based on the KOs and grouping of the KOs is avoided.

```{r}
#| label: fig-n-metabolism

p3 <- kos %>% 
  filter(metabolism == "N metabolism") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(pathway),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = kos %>% 
                     filter(metabolism == "N metabolism") %>% 
                     select(pathway) %>% distinct() %>% 
                     mutate(pathway = if_else(grepl("chlMax", pathway), "", pathway)) %>%
                     pull(pathway) %>% rev()
                     ) +
  labs(x = "", y = "") + 
  theme_tidy(fontsize = 5) + 
  theme(
    axis.ticks.y = element_blank(),
    aspect.ratio = 1.0
  )

p3
```


```{r}
#| label: fig-photosystems

p4 <- kos %>% 
  filter(metabolism == "Photosystems and ATP") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(pathway),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = kos %>% 
                     filter(metabolism == "Photosystems and ATP") %>% 
                     select(pathway) %>% distinct() %>% 
                     mutate(pathway = if_else(grepl("chlMax", pathway), "", pathway)) %>%
                     pull(pathway) %>% rev()
                     ) +
  labs(x = "", y = "") + 
  theme_tidy(fontsize = 5)

p4
```


```{r}
#| label: fig-s-metabolism

p5 <- kos %>% 
  filter(metabolism == "S metabolism") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(pathway),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = kos %>% 
                     filter(metabolism == "S metabolism") %>% 
                     select(pathway) %>% distinct() %>% 
                     mutate(pathway = if_else(grepl("chlMax", pathway), "", pathway)) %>%
                     pull(pathway) %>% rev()
                     ) +
  labs(x = "", y = "") + 
  theme_tidy(fontsize = 5) +
  theme(
    axis.ticks.y = element_blank(),
    aspect.ratio = 1.0
  )

p5
```


```{r}
#| fig-p-metabolism

p6 <- kos %>% 
  filter(metabolism == "P metabolism") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(pathway),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = kos %>% 
                     filter(metabolism == "P metabolism") %>% 
                     select(pathway) %>% distinct() %>% 
                     mutate(pathway = if_else(grepl("chlMax", pathway), "", pathway)) %>%
                     pull(pathway) %>% rev()
                     ) +
  labs(x = "", y = "", fill = "") +
  theme_tidy(fontsize = 5) +
  theme(
    axis.ticks.y = element_blank(),
    legend.key.width = unit(3.5, "mm"),
    legend.key.height = unit(4, "mm"),
    aspect.ratio = 1.0
  )

p6
```



@fig-elements shows the tpm for each representative KO summed over all the samples (n = 20). N transformation, central C metabolism, and photosynthesis plus aerobic respiration. Udh uronate dehydrogenase, edd phosphogluconate dehydratase, icd isocitrate dehydrogenase, ICL isocitrate lyase, G6PD glucose-6-phosphate 1-dehydrogenase, PRPP phosphoribosyl pyrophosphate, RPPK ribose-phosphate pyrophosphokinase, ED: Entner–Doudoroff. Note how the transcripts involved in carbon fixation are predominantly represented by the Streptomycetales (Streptomyces sp.) and the Burkholderiales (Nitrosomonas sp.) while only few transcripts were detected affiliated with cyanobacteria. The oxidation from ammonia to nitrite (nitrification) is performed by Nitrosopumilus sp. while the further oxidation from nitrite to nitrate (K00370) is done by a representative from the family Nitrospinaceae.

```{r}
#| label: fig-elements
#| fig-cap: KO plot

(p1 + p3 + p6) + 
  plot_annotation(tag_levels = list(c('c', 'd', 'e'))) + 
  plot_layout(guides = "collect") & guides(fill = guide_legend(reverse = TRUE)) &
  theme(
    legend.key.height = unit(3.5, "mm"), legend.key.width = unit(2.5, "mm"),
    plot.margin = margin(t = 10),
    plot.tag.location = "panel",
    plot.tag.position = c(0.93, 0.95),
    axis.ticks.y = element_blank(),
    legend.position = "bottom", legend.title = element_blank(),
    legend.margin = margin(t = -10), legend.text = element_text(margin = margin(l = 2))
    )
```


```{r}
ggsave("figures/ko-modules.pdf", width = 180, height = 140, units = "mm")
```


# Linking metat to morphology (SEM)


```{r}
#| label: KO filter morphology

kos <- kofam %>%
  separate_longer_delim(ko, delim = ",") %>%
  mutate(label = case_when(
    # Pilus system: Pilus assembly proteins
    ko %in% c('K02650','K02651','K02652','K02653','K02654','K02656','K02662','K02663','K02664','K02665','K02666',
              'K02671','K02672','K02673','K02674','K02675','K02655','K23985','K02676') ~ 'Pilli (pilAVCDEFMNOPQVWYZ): External appendages\nand membrane extensions',
    ko %in% c('K02278','K02279','K02280','K02281','K02282','K02283') ~ "Pilli (cpaBCDEF): External appendages\nand membrane extensions",
    # EPS biosynthesis:
    ko %in% c("K13693","K13683","K03818","K16710","K16703","K03208","K03606") ~ "wcaAEFKLIJ: EPS biosynthesis",
    ko %in% c("K00903","K19421","K19422") ~ "epsBCD: EPS biosynthesis",
    ko %in% c("K00703","K00975","K00688") ~ "glgACP: EPS biosynthesis", # Biofilm formation E. coli (map02026)
    ko == 'K03563' ~ 'csrA: EPS biosynthesis', # Biofilm formation E. coli (map02026)
    # Flagellar system
    ko %in% c("K02410","K02416","K02417") ~ "Flagella (fliACDEFGKLMNS): External appendages\nand membrane extensions",
    # M, S, P and L rings
    ko %in% c("K02409","K02415") ~ "Flagella (fliACDEFGKLMNS): External appendages\nand membrane extensions",
    ko %in% c("K02394","K02386","K02393") ~ "Flagella (flgABCDEFGHIJKL): External appendages\nand membrane extensions",
    # Rod and hook
    ko %in% c('K02408','K02414') ~ "Flagella (fliACDEFGKLMNS): External appendages\nand membrane extensions",
    ko %in% c('K02387','K02388','K02389','K02391','K02392','K02395','K02390','K02396','K02397','K02399') ~ "Flagella (flgABCDEFGHIJKL): External appendages\nand membrane extensions",
    # Filament
    ko %in% c('K02406','K02407','K02422') ~ "Flagella (fliACDEFGKLMNS): External appendages\nand membrane extensions",
    # Stator
    ko == 'K02405' ~ "Flagella (fliACDEFGKLMNS): External appendages\nand membrane extensions",
    ko %in% c('K02556','K02557') ~ 'Flagella (motAB): External appendages\nand membrane extensions',
    # Nanotube
    ko %in% c("K02418","K02419","K02420") ~ 'Nanotubes (fliOPQ): External appendages\nand membrane extensions',
    ko == 'K02400' ~ 'Nanotubes (flhA): External appendages\nand membrane extensions',
    ko == 'K09769' ~ 'Nanotubes (ymdB): External appendages\nand membrane extensions',
    # Stalk
    ko %in% c("K07657","K07636") ~ "Stalk (phoBR): External appendages\nand membrane extensions",
    ko %in% c("K02038","K02036","K02037") ~ "Stalk (pstABC): External appendages\nand membrane extensions",
    # Vesicle
    ko == 'K04754' ~ 'vacJ: Vesicle formation',
    ko == 'K03640' ~ 'pal: Vesicle formation',
    ko == 'K04771' ~ 'degP: Vesicle formation',
    ko %in% c('K09476','K11934') ~ 'ompFX: Vesicle formation',
    ko %in% c('K03646','K03641','K03560','K03562') ~ 'tolABRQ: Vesicle formation',
    ko == "K06078" ~ 'lpp: Vesicle formation',
    ko == "K20257" ~ 'pqsE: Vesicle formation',
  )) %>%
  filter(!is.na(label)) %>%
  separate_wider_delim(label, delim = ": ", names = c("label", "category")) %>%
  inner_join(tax, by = "orf") %>% 
  inner_join(tpm, by = "orf") %>% 
  inner_join(meta, by = "sample") %>%
  replace_na(list("order" = "Other")) %>%
  mutate(order = if_else(order %in% names(cols.order), order, "Other")) %>%
  group_by(layer, label, order, category) %>% summarise(tpm = sum(tpm), .groups = "drop") %>%
  mutate(order = factor(order, names(cols.order))) %>%
  mutate(i = if_else(
     layer == "10m", true = label, false = paste(label, layer)
   ))
```


```{r}
#| label: color palette morphology

kos %>%
  group_by(order) %>%
  summarise(n = sum(tpm)) %>% filter(!is.na(order)) %>%
  arrange(desc(n))
```




```{r}
cols.order <- c( # Sorted on abundance
  "Pseudomonadales" = "#C7B19C",
  "Pelagibacterales" = "#972D15",
  "SAR324" = "#D69C4E",
  "Marinisomatales" = "#046C9A",
  "Puniceispirillales_A" = "#9986A5",
  "PS1" = "#FAEFD1",
  "Rhodobacterales" = "#00A08A",
  "SAR86" = "#D5D5D3",
  "Burkholderiales" = "#78B7C5",
  "UBA8366" = "#A2A475",
  "Flavobacteriales" = "#D8B70A",
  "Other" = "#D5D5D3"
) %>% rev()

lab <- kos %>%
  filter(category == "EPS biosynthesis") %>%
  arrange(i) %>%
  select(i) %>%
  distinct() %>%
  mutate(i = if_else(grepl("chlMax", i), "", i)) %>%
  pull(i) %>% rev()

p1 <- kos %>%
  filter(category == "EPS biosynthesis") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(i),
    fill = order
    )) +
  geom_col(width = 0.85, show.legend = TRUE) +
  scale_fill_manual(values = cols.order, drop = FALSE) +
  labs(x = "Relative abundance (tpm)", y = "") + 
  theme_tidy(legend = "right", fontsize = 5) +
  scale_y_discrete(labels = lab) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme(
    legend.title = element_blank(),
    axis.ticks.y = element_blank(),
    legend.text = element_text(margin = margin(l = 2)),
    legend.key.height = unit(3.5, "mm"), legend.key.width = unit(2.5, "mm"),

  )

p1
```

```{r}
lab <- kos %>%
  filter(category == "Vesicle formation") %>%
  arrange(i) %>%
  select(i) %>%
  distinct() %>%
  mutate(i = if_else(grepl("chlMax", i), "", i)) %>%
  pull(i) %>% rev()

p2 <- kos %>%
  filter(category == "Vesicle formation") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(i),
    fill = order
    )) +
  geom_col(width = 0.85, show.legend = TRUE) +
  scale_fill_manual(values = cols.order, drop = FALSE) +
  labs(x = "Relative abundance (tpm)", y = "") + 
  scale_y_discrete(labels = lab) +
  theme_tidy(legend = "right", fontsize = 5) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme(
    legend.title = element_blank(),
    axis.ticks.y = element_blank(),
    legend.text = element_text(margin = margin(l = 2)),
    legend.key.height = unit(3.5, "mm"), legend.key.width = unit(2.5, "mm"),

  )

p2
```

```{r}
lab <- kos %>%
  filter(category == "External appendages\nand membrane extensions") %>%
  arrange(i) %>%
  select(i) %>%
  distinct() %>%
  mutate(i = if_else(grepl("chlMax", i), "", i)) %>%
  pull(i) %>% rev()

p3 <- kos %>%
  filter(category == "External appendages\nand membrane extensions") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(i),
    fill = order
    )) +
  geom_col(width = 0.85, show.legend = TRUE) +
  scale_fill_manual(values = cols.order, drop = FALSE) +
  labs(x = "Relative abundance (tpm)", y = "") + 
  theme_tidy(legend = "right", fontsize = 5) +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_y_discrete(labels = lab) +
  theme(
    legend.title = element_blank(),
    axis.ticks.y = element_blank(),
    legend.text = element_text(margin = margin(l = 2)),
    legend.key.height = unit(3.5, "mm"), legend.key.width = unit(2.5, "mm"),

  )

p3
```


```{r}
ggsave(filename = "figures/morph-draft.pdf", width = 180, height = 120, units = "mm")
```



# Osmoregulation

```{r}
d <- read_tsv("data/genelist_maintenance.tsv", show_col_types = F) %>%
  rename(process = subcat) %>%
  mutate(subcategory = case_when(
    grepl("SSBR", process) ~ "SSBR",
    grepl("BER", process) ~ "BER",
    grepl("MMR", process) ~ "MMR",
    grepl("DSBR", process) ~ "DSBR",
    grepl("NER", process) ~ "NER",
    grepl("CRISPR-Cas", process) ~ "CRISPR-Cas",
    grepl("R-M system", process) ~ "R-M system",
    grepl("TA system", process) ~ "TA system",
    grepl("DNA phosphothiolation system", process) ~ "DNA P system",
    .default = process
    )) %>%
  mutate(category = case_when(
    category == "defense" ~ "Defense system",
    category == "osmoregulation" ~ "Osmoregulation",
    .default = category
  ))
```



```{r}
d <- d %>%
  left_join(kofam, by = "ko", relationship = "many-to-many") %>%
  inner_join(tax, by = "orf") %>%
  inner_join(tpm, by = "orf") %>%
  inner_join(meta, by = "sample") %>%
  mutate(layer = factor(layer, c("10m", "chlMax"))) %>%
  mutate(i = if_else(
     layer == "10m", true = subcategory, paste(subcategory, layer)
   )) %>%
  mutate(order = if_else(is.na(order), "Other", order)) %>%
  mutate(order = if_else(!order %in% names(cols.order), "Other", order)) %>%
  mutate(order = factor(order, levels = names(cols.order))) %>%
  group_by(layer, order, i, category) %>% summarise(tpm = sum(tpm), .groups = "drop")
```

```{r}
cols.order <- c( # Sorted on abundance
  "Pseudomonadales" = "#C7B19C",
  "Pelagibacterales" = "#972D15",
  "Poseidoniales" = "#D69C4E",
  "Marinisomatales" = "#046C9A",
  "Puniceispirillales_A" = "#9986A5",
  "PS1" = "#FAEFD1",
  "Rhodobacterales" = "#00A08A",
  "SAR86" = "#1B5331",
  "CACEW01" = "#78B7C5",
  "UBA8366" = "#A2A475",
  "Flavobacteriales" = "#D8B70A",
  "Other" = "#D5D5D3"
) %>% rev()
```


```{r}
#| label: fig-defense
#| fig-cap: KEGG orthologs involved in osmoregulation or the regulation of osmoregulation. Each bar shows the summed tpm (n = 10) for the depth layers. Showing the ten most relevant orders while grouping low-abundant lineages and transcripts without taxonomy as "Other". The x-axis was transformed with a square-root to also visualize KOs with a low tpm.

lab <- d %>%
  filter(category == "Defense system") %>%
  arrange(i) %>%
  select(i) %>%
  distinct() %>%
  mutate(i = if_else(grepl("chlMax", i), "", i)) %>%
  pull(i) %>% rev()

px <- d %>%
  filter(category == "Defense system") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(i),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE, width = 0.8) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = lab) +
  labs(x = "Relative abundance (tpm)", y = "", fill = "") + 
  theme_tidy(fontsize = 5, legend = "right") +
  guides(fill = guide_legend(reverse = T)) +
  theme(
     axis.ticks.y = element_blank(),
     legend.key.height = unit(3.5, "mm"),
     aspect.ratio = 1.4,
     panel.border = element_rect(fill = "transparent", linewidth = 0.75)
   )

px
```

```{r}
#| label: fig-defense
#| fig-cap: KEGG orthologs involved in osmoregulation or the regulation of osmoregulation. Each bar shows the summed tpm (n = 10) for the depth layers. Showing the ten most relevant orders while grouping low-abundant lineages and transcripts without taxonomy as "Other". The x-axis was transformed with a square-root to also visualize KOs with a low tpm.

lab <- d %>%
  filter(category == "Osmoregulation") %>%
  arrange(i) %>%
  select(i) %>%
  distinct() %>%
  mutate(i = if_else(grepl("chlMax", i), "", i)) %>%
  pull(i) %>% rev()

py <- d %>%
  filter(category == "Osmoregulation") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(i),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE, width = 0.8) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = lab) +
  labs(x = "Relative abundance (tpm)", y = "", fill = "") + 
  theme_tidy(fontsize = 5, legend = "right") +
  guides(fill = guide_legend(reverse = T)) +
  theme(
     axis.ticks.y = element_blank(),
     legend.key.height = unit(3.5, "mm"),
     aspect.ratio = 1.4,
     panel.border = element_rect(fill = "transparent", linewidth = 0.75)
   )

py
```

```{r}
#| label: fig-defense
#| fig-cap: KEGG orthologs involved in osmoregulation or the regulation of osmoregulation. Each bar shows the summed tpm (n = 10) for the depth layers. Showing the ten most relevant orders while grouping low-abundant lineages and transcripts without taxonomy as "Other". The x-axis was transformed with a square-root to also visualize KOs with a low tpm.

lab <- d %>%
  filter(category == "DNA repair and recombination") %>%
  arrange(i) %>%
  select(i) %>%
  distinct() %>%
  mutate(i = if_else(grepl("chlMax", i), "", i)) %>%
  pull(i) %>% rev()

pz <- d %>%
  filter(category == "DNA repair and recombination") %>%
  ggplot(aes(
    x = tpm, 
    y = fct_rev(i),
    fill = fct_relevel(order, names(cols.order))
    )) +
  geom_col(show.legend = TRUE, width = 0.8) + 
  scale_fill_manual(values = cols.order, drop = FALSE) +
  scale_y_discrete(labels = lab) +
  labs(x = "Relative abundance (tpm)", y = "", fill = "") + 
  theme_tidy(fontsize = 5, legend = "right") +
  guides(fill = guide_legend(reverse = T)) +
  theme(
     axis.ticks.y = element_blank(),
     legend.key.height = unit(3.5, "mm"),
     aspect.ratio = 1.4,
     panel.border = element_rect(fill = "transparent", linewidth = 0.75)
   )

pz
```

```{r}
#| label: fig-basal
#| fig-cap: KO plot

(p1 + p2 + p3) / (px + py + pz) + 
  plot_annotation(tag_levels = list(c('EPS biosynthesis', 'Vesicle formation', 'External appendages\nand membrane extensions','Defense system', 'Osmoregulation', 'DNA repair and recombination'))) + 
  plot_layout(guides = "collect") & guides(fill = guide_legend(reverse = TRUE)) &
  theme(
    legend.key.height = unit(3.5, "mm"), legend.key.width = unit(2.5, "mm"),
    plot.margin = margin(t = 10),
    plot.tag.location = "panel",
    plot.tag.position = c(0.5, 1.05), plot.tag = element_text(size = 5, face = "plain"),
    axis.ticks.y = element_blank(),
    legend.position = "right", legend.title = element_blank(),
    legend.margin = margin(t = -10), legend.text = element_text(margin = margin(l = 2))
    )
```






```{r}
ggsave(filename = "figures/maintenance.pdf", width = 180, height = 140, units = "mm")
```


# Differential gene expression using edgeR

This analysis helps to identify differentially expressed genes between the 10 m depth and the chlorophyll max depth. Station 5 is not included in the analysis as it introduces too much noise.

```{r}
library(edgeR)
i <- meta %>% filter(!station %in% c("5", "56")) %>% pull(sample)

m <- tpm %>%
  filter(sample %in% i) %>% # Leave out station 5 for the moment
  select(-tpm) %>%
  arrange(sample) %>%
  pivot_wider(values_from = "count", names_from = "sample", values_fill = 0) %>%
  column_to_rownames("orf") # Create a numeric matrix

dge <- tpm %>%
  filter(sample %in% i) %>% # Leave out station 5 for the moment
  select(-tpm) %>%
  inner_join(eggnog, by = "orf") %>% separate_longer_delim(ko, delim = ",") %>%
  group_by(sample, ko) %>% summarise(count = sum(count), .groups = "drop") %>%
  arrange(sample) %>%
  pivot_wider(values_from = "count", names_from = "sample", values_fill = 0) %>%
  column_to_rownames("ko") %>%
  DGEList(group = meta %>% filter(sample %in% i) %>% pull(layer), lib.size = colSums(m))
```

The chunk below is to create the design matrix, make the contract, and run the actual analysis using the functions glmQLfit() and glmQLFTest()

```{r}
design.mat <- model.matrix(~ 0 + meta %>% filter(sample %in% i) %>% arrange(sample) %>% pull(layer))
rownames(design.mat) <- colnames(m)
colnames(design.mat) <- c("shallow", "deep")
contrast <- makeContrasts(layer = shallow-deep, levels = design.mat)

dge <- dge %>%
  calcNormFactors() %>%
  estimateDisp(design.mat) %>%
  glmQLFit(design.mat, robust = TRUE)

qlf <- glmQLFTest(dge, contrast = contrast)
```

The p-values are adjusted according to the false discovery rate, and the DE genes are joined with the functional and taxonomic annotation.

```{r}
de <- topTags(qlf, adjust.method = "fdr", n = Inf) %>%
  as.data.frame() %>%
  rownames_to_column("ko") %>% rename_with(tolower) %>%
  mutate(expression = case_when(
    fdr < 0.05 & logfc > 0 ~ "up",
    fdr < 0.05 & logfc < 0 ~ "down",
    .default = "non-de"
  )) %>%
  inner_join(eggnog, by = "ko") %>%
  filter(expression != "non-de") %>% tibble()

remove(m, dge, qlf, contrast, design.mat)
```

As the contrast was shallow vs deep, upregulated means significantly higher in the shallow samples while downregulated means higher in the deeper samples. There are `r de %>% filter(expression == "up") %>% pull(ko) %>% unique() %>% length()` KOs up regulated and `r de %>% filter(expression == "down") %>% pull(ko) %>% unique() %>% length()` KOs down regulated. `r de %>% filter(expression == "non-de") %>% pull(ko) %>% unique() %>% length()` KOs were not differential expressed.

```{r}
cols.genus <- c(
  "Pseudothioglobus" = "#C7B19C",
  "AG-430-B22" = "#A2A475",
  "Nitrosopumilus" = "#972D15",
  "Pelagibacter" = "#D69C4E",
  "Alteromonadaceae uncl." = "#FAEFD1",
  "ASP10-02a" = "#00A08A",
  "Sneathiella" = "#D8B70A",
  "AG-337-I02" = "#1B5331",
  "AG-430-B22" = "#D5D5D3",
  "Other" = "#046C9A"
)
  
cols.order <- c( # Sorted on abundance
  "Rhizobiales_B" = "#D69C4E",
  "Pelagibacterales" = "#972D15",
  "Rhodobacterales" = "#D8B70A",
  "Rhodospirillales" = "#00A08A",
  "Pseudomonadales" = "#C7B19C",
  "Rhizobiales" = "#D5D5D3",
  "HIMB59" = "#1B5331",
  "Nitrososphaerales" = "#A2A475",
  "PS1" = "#FAEFD1",
  "Other" = "#046C9A"
)
```

The 25 KOs with the lowest (up in deeper samples) and 25 with the highest log fold-change are selected. Six of these KOs are represented in the gene list, the other 19 will be manually categorized. The number of 25 was arbitrary chosen. Many of these KOs are affiliated with the archaeon Nitrosopumilus that oxidizes ammonium and is capable of autotrophic growth using the hydroxypropionate/hydroxybutyrate cycle that is more energy efficient than other aerobic C fixation pathways, including the Calvin cycle (source: https://doi.org/10.1073/pnas.1402028111). The fact that C fixation KOs affiliated with this genus are abundantly detected strongly suggests autotrophic growth, possibly due to a low concentration of organic carbon.

```{r}
top.de <- de %>%
  select(ko, logfc) %>% distinct() %>%
  slice_max(logfc, n = 20)

top.de <- de %>%
  select(ko, logfc) %>% distinct() %>%
  slice_min(logfc, n = 20) %>%
  rbind(top.de)
```



```{r}
top.de <- de %>%
  select(ko, logfc) %>% distinct() %>%
  mutate(pathway = case_when(
    ko == "K01638" ~ "Glycolate utilization (glcB)",
    ko == "K03320" ~ "Ammonium transporter (amtAB)",
    ko == "K00372" ~ "Assimilatory nitrate reductase (nasA)",
    ko == "K00366" ~ "Ferredoxin-nitrite reductase: nitrate-ammonia (nirA)",
    ko == "K00362" ~ "Nitrite reductase (nirB)",
    ko == "K00363" ~ "Nitrite reductase (nirD)",
    ko == "K07112" ~ "Sulphur transport",
    ko == "K02256" ~ "Cytochrome C and quinol oxidase",
    ko == "K17225" ~ "Sulfur oxidation soxC",
    ko == "K17227" ~ "Sulfur oxidation soxZ",
    ko == "K00394" ~ "Dissimilatory sulfate reduction (aprA)",
    ko == "K00381" ~ "Assimilatory sulfate reduction (cysI)",
    ko == "K17223" ~ "Sulfur oxidation SoxX",
    ko == "K00387" ~ "Sulfite oxidation (SUOX)",
    ko == "K03518" ~ "CO oxidation (coxS)",
    ko == "K03519" ~ "CO oxidation (coxM, cutM)",
    ko == "K03520" ~ "CO oxidation (coxL, cutL)",
    ko == "K00363" ~ "Dissimilatory nitrite reduction (nirD)",
    ko == "K02044" ~ "Phosphonate, periplasmic substrate-binding protein",
    ko == "K02027" ~ "Carbohydrate uptake (transported)",
    ko == "K07463" ~ "Archaea-specific exonuclease",
    ko == "K15016" ~ "Hydroxypropionate-hydroxybutylate cycle",
    ko == "K21817" ~ "Synthesis of the retinal cofactor of bacteriorhodopsin",
    ko == "K07725" ~ "CRISPR-associated protein",
    ko == "K00123" ~ "Formate dehydrogenase (fdhF)",
    ko == "K05973" ~ "Poly(3-hydroxybutyrate) depolymerase (phaZ)"
  )) %>% filter(!is.na(pathway))
```

```{r}
#| fig-cap: Differentially expressed orthologous genes (KOs) between 10 m depth and chlorophyll max (between 20 and 45 m depth). A positive fold change indicates the KO was detected in significant higher numbers at 10 m depth, while a negative value indicates higher numbers at chlorophyll max.
#| label: fig-de

top.de %>%
  mutate(ko = paste(ko, pathway, sep = ": ")) %>%
  ggplot(aes(x = logfc, 
             y = ko, 
             #fill = order
             )
         ) +
  geom_col() + geom_vline(xintercept = 0, linewidth = unit(0.5, "mm"), color = "grey") +
  labs(x = expression("Log"["2"]*" fold change"), y = "", fill = "") +
  #scale_fill_manual(values = cols.order) + theme_tidy(fontsize = 6, legend = "right") +
  theme(
    aspect.ratio = 1.5,
    axis.ticks.y = element_blank(),
    legend.key.height = unit(3.5, "mm"),
    legend.key.spacing.y = unit(0, "mm"),
    panel.border = element_rect(fill = "transparent", linewidth = unit(0.5, "mm"))
  )
```

