---
title: "Synoptic Arctic Survey 2021 - metatranscriptomics"
author: "geweaa"
bibliography: references.bib
csl: diabetologia.csl
date: now
format: html
toc: true
toc_float: true
toc_depth: 3
number_sections: true
theme: sandstone
mainfont: Helvetica
monofont: AndaleMono
fontsize: "9"
highlight: textmate
df_print: paged
code-fold: true
code-summary: "Expand code"
self_contained: true
embed-resources: true
keep_md: false
encoding: "UTF-8"
other-links:
      - text: GitHub repo (private for now)
        href: https://github.com/geweaa/sas-metat
      - text: Nf-core/metatdenovo
        href: https://nf-co.re/metatdenovo/1.0.1/
      - text: Nature guidelines on figures
        href: https://research-figure-guide.nature.com/figures/preparing-figures-our-specifications/
      - text: The English Bookshop
        href: https://bookshop.se
---


```{r libraries, include=FALSE}
library(tidyverse)
source("../Protocols/theme.R")
knitr::opts_chunk$set(cache = TRUE)
```


The raw data was procesesd using nf-core/metatdenovo (v1.0.1): https://nf-co.re/metatdenovo/1.0.1/ 
The following command was used to start the pipeline: nextflow run nf-core/metatdenovo -r dev -profile uppmax -params-file params.yaml
The file params.yaml contained the following relevant parameters: 
    skip_kofamscan: TRUE 
    orf_caller: 'prokka' 
    eukulele_db: 'gtdb-r220'
    input: 'input.csv'
    project: naiss2024-22-406

Apart from this, the pipeline was run on default parameters. GTDB release 220 was used for taxonomic annotation.

```{r read the taxonomy, message=FALSE}
tax <- read_tsv("data/megahit.prokka.gtdb_taxonomy.tsv.gz", show_col_types = F)
```


```{r read the meta-t files}
tpm <- read_tsv("data/megahit.prokka.counts.tsv.gz", show_col_types = F) %>% 
  select(orf, sample, count, tpm) %>%
  mutate(sample = gsub("R", ".", sample)) %>%
  mutate(sample = factor(sample, levels = c(
    "S1.1", "S1.2", "S2.1", "S2.2", "S3.1", "S3.2", "S4.1", "S4.2", "S5.1", "S6.1", 
    "S5.2", "S6.2", "S7.1", "S7.2", "S8.1", "S8.2", "S9.1", "S9.2", "S10.1", "S10.2"
  )))

meta <- read_tsv("data/sample_key.tsv", show_col_types = FALSE) %>%
  rename_with(tolower) %>%
  mutate(station = as.character(station)) %>%
  mutate(sample = gsub("R", ".", sample)) %>%
  select(-`user id`, -`>=q30(%)`) %>%
  mutate(sample = factor(sample, levels = c(
    "S1.1", "S1.2", "S2.1", "S2.2", "S3.1", "S3.2", "S4.1", "S4.2", "S5.1", "S6.1", 
    "S5.2", "S6.2", "S7.1", "S7.2", "S8.1", "S8.2", "S9.1", "S9.2", "S10.1", "S10.2"
  ))) %>%
  mutate(depth = factor(depth, levels = c("chlMax", "10m")))
```

In total, there are `r nrow(meta)` metatranscriptomes from five stations. At every station, samples were taken in duplicates from 10 m depth and from chl. max. On average, each sample has `r round(mean(meta$mreads), digits = 1)` million reads (sd `r sd(meta$mreads) %>% round(digits = 1)`, n = 20). Off all the transcripts, 53.0 % was functionally annotated (Eggnog mapper), 36.5 % was functionally annotated including a KO reference, 45.7 % of the transcripts was annotated on phylum level, 45.7 % on order level, 44.8 % on genus level, and 32.7 % on species level. 

25.2 % of the transcript were functionally annotated with Prokka. A summary of the bioinformatic pipeline will soon be added to this introduction.

```{r read the functional annotations}
eggnog <- read_tsv("data/megahit.prokka.emapper.tsv.gz", show_col_types = F) %>%
  rename_with(tolower) %>%
  mutate(kegg_ko = gsub("ko:", "", kegg_ko))
```



```{r define colors}
cols.phylum <- c( # Sorted on abundance
  "Proteobacteria" = "#C7B19C",
  "Thermoproteota" = "#A2A475",
  "Thermoplasmatota" = "#972D15",
  "Cyanobacteria" = "#D69C4E",
  "Bacteroidota" = "#FAEFD1",
  "Marinisomatota" = "#00A08A",
  "Chloroflexota" = "#D8B70A",
  "Actinobacteriota" = "#1B5331",
  "Other" = "#046C9A",
  "Unidentified" = "#D5D5D3"
)

cols.order <- c( # Sorted on abundance
  "Pseudomonadales" = "#C7B19C",
  "Nitrososphaerales" = "#A2A475",
  "Pelagibacterales" = "#972D15",
  "Poseidoniales" = "#D69C4E",
  "PS1" = "#FAEFD1",
  "Cyanobacteriales" = "#00A08A",
  "Flavobacteriales" = "#D8B70A",
  "SAR86" = "#1B5331",
  "Gloeomargaritales" = "#D5D5D3",
  "Other" = "#046C9A"
)

cols.genus <- c( # Sorted on abundance
  "Pseudothioglobus" = "#C7B19C",
  "Ketobacter" = "#A2A475",
  "Nitrosopumilus" = "#972D15",
  "Pelagibacter" = "#D69C4E",
  "HTCC2207" = "#FAEFD1",
  "ASP10-02a" = "#00A08A",
  "Pelagibacter_A" = "#D8B70A",
  "JACYLR01" = "#1B5331",
  "MGIIb-O3" = "#046C9A",
  "AAA536-G10" = "#D5D5D3"
)

cols.station <- c(
   "5" = "#046C9A",
  "26" = "#446455",
  "42" = "#FDD262",
  "53" = "#D3DDDC",
  "56" = "#C7B19C"
)
```


# Data exploration

@fig-domain depicts the ratio between bacteria and archaea for both depth layers. Note how the abundance of archaea is slightly higher in most of the deeper samples (chlorophyll max depth). 

```{r}
#| fig-cap: Community overview on domain level, removing transcripts without a taxonomic annotation.
#| label: fig-domain
#| out-width: 100%
#| fig-width: 4.75

tpm %>%
  inner_join(tax, by = "orf") %>%
  inner_join(meta, by = "sample") %>%
  group_by(sample, domain) %>% summarise(tpm = sum(tpm), .groups = "drop") %>%
  filter(domain %in% c("Bacteria", "Archaea")) %>%
  mutate(sample = factor(sample, levels = c(
    "S1.1", "S1.2", "S3.1", "S3.2", "S5.1", "S6.1", "S7.1", "S7.2", "S9.1", "S9.2", 
    "S2.1", "S2.2", "S4.1", "S4.2", "S5.2", "S6.2", "S8.1", "S8.2", "S10.1", "S10.2"
  ))) %>%
  ggplot(aes(
    x = sample,
    y = tpm,
    fill = domain
  )) +
  geom_col() + theme_tidy(ratio = 0.7) +
  scale_y_continuous(breaks = c(0, 2e5, 4e5), labels = c(0, 2, 4)) +
  labs(y = expression('Relative abundance (tpm '%*% '10'^'5'*')')) + 
  scale_fill_manual(values = c("Bacteria" = "#C7B19C", "Archaea" = "#A2A475")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme(
    legend.title = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  ) +
  annotate("segment", x =  0.55, xend =  10.45, y = -15000, yend = -15000) +
  annotate("segment", x =  10.55, xend = 20.45, y = -15000, yend = -15000) +
  annotate("label", x = 5, y = -17000, label = "10 m", label.size = 0, size = 7/.pt) +
  annotate("label", x = 15, y = -17000, label = "Chl. max", label.size = 0, size = 7/.pt)
```



```{r most abundant phyla for plotting}
#| tbl-cap: The eight most abundant phyla of all samples, based on summed TPM. Most are bacterial phyla, except for the archaeal phyla Thermoproteota and Thermoplasmatota. 

t <- tpm %>%
  inner_join(tax, by = "orf") %>%
  group_by(phylum, sample) %>%
  # Sum the abundance of each phylum within a sample
  summarise(tpm = sum(tpm), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean.tpm = sum(tpm), .groups = 'drop') %>%
  filter(!is.na(phylum)) %>%
  top_n(8, mean.tpm)

t %>% arrange(desc(mean.tpm)) %>% knitr::kable()
```


```{r prepare the taxonomy, include=FALSE}
taxref <- tax %>%
  left_join(t %>% transmute(phylum, topphylum = phylum), by = "phylum") %>%
  replace_na(list("topphylum" = "Other"))
```


## Community composition - phylum

@fig-phylum explores the data by plotting the most abundant phyla over all samples, regardless if the RNA transcript was functionally annotated or not. Note that the taxonomy is outdated and needs to be updated (e.g. Pseudomonadota instead of Proteobacteria). 87.7 % of the annotated ORFs are affiliated with bacteria, while 13.1 % is affiliated with Archaea. 


```{r}
#| label: fig-phylum
#| fig-cap: Community composition of all RNA transcripts on phylum level. The phyla are sorted according to relative abundance while grouping low-abundant phyla as 'Other' and transcripts without any taxonomic annotation as 'Unidentified'
#| out-width: 100%

tpm %>%
  left_join(taxref, by = "orf") %>%
  replace_na(list("topphylum" = "Unidentified")) %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(sample, topphylum) %>% 
  summarise(tpm = sum(tpm), .groups = 'drop') %>%
  # Call the plot
  ggplot(aes(x = sample, 
             y = tpm, 
             fill = fct_relevel(topphylum, names(cols.phylum) %>% rev())
             )
         ) +
  labs(x = '', y = expression('Relative abundance (tpm '%*% '10'^'6'*')'), fill = "Phylum") +
  geom_col() +
  scale_y_continuous(breaks = c(0, 2e5, 4e5, 6e5, 8e5, 1e6),
                labels = c("0.0", "0.2", "0.4", "0.6", "0.8", "1.0")
                ) +
  scale_fill_manual(values = cols.phylum) +
  theme_tidy(ratio = 0.5) + 
  theme(
    axis.ticks.x = element_blank(),
    legend.key.height = unit(3.5, "mm"),
    legend.margin = margin(),
    panel.border = element_rect(fill = "transparent", linewidth = 0.5),
    axis.text.x = element_blank()
    ) +
  annotate("segment", x =  0.55, xend =  4.45, y = -25000, yend = -25000) +
  annotate("segment", x =  4.55, xend =  8.45, y = -25000, yend = -25000) +
  annotate("segment", x =  8.55, xend = 12.45, y = -25000, yend = -25000) +
  annotate("segment", x = 12.55, xend = 16.45, y = -25000, yend = -25000) +
  annotate("segment", x = 16.55, xend = 20.45, y = -25000, yend = -25000) +
  annotate("label", x = 2.5, y = -25000, label = "Station 5", label.size = 0, size = 7/.pt) +
  annotate("label", x = 6.5, y = -25000, label = "26", label.size = 0, size = 7/.pt) +
  annotate("label", x =10.5, y = -25000, label = "42", label.size = 0, size = 7/.pt) +
  annotate("label", x =14.5, y = -25000, label = "53", label.size = 0, size = 7/.pt) +
  annotate("label", x =18.5, y = -25000, label = "56", label.size = 0, size = 7/.pt)
```


## Community composition - order


```{r most abundant orders for plotting}
#| tbl-cap: The eight most abundant orders, after summing the TPM of all the samples.

t <- tpm %>%
  inner_join(tax, by = "orf") %>%
  group_by(order, sample) %>%
  # Sum the abundance of each phylum within a sample
  summarise(tpm = sum(tpm), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean.tpm = sum(tpm), .groups = 'drop') %>%
  filter(!is.na(order)) %>%
  top_n(8, mean.tpm)

t %>% arrange(desc(mean.tpm)) %>% knitr::kable()
```


```{r prepare the taxonomy - order, include=FALSE}
taxref <- tax %>%
  left_join(t %>% transmute(order, topphylum = order), by = "order") %>%
  replace_na(list("topphylum" = "Other"))
```


@fig-order explores the data by plotting the most abundant orders over all samples, regardless if the RNA transcript was functionally annotated or not. Note that transcripts without a taxonomic annotation (the unidentified fraction in @fig-phylum) are not included in this figure. It clearly shows how the Nitrosphaerales are more abundant in the deeper layer.


```{r}
#| label: fig-order
#| fig-cap: Community composition of all RNA transcripts on order level. The orders are sorted according to relative abundance while grouping low-abundant orders as 'Other'. Transcripts without taxonomic annotation are not included. For each station, the two bars left of the label were sampled at 10 m depth, while the two bars on the right were sampled at chlorophyll max (between 20 and 45 m depth)
#| out-width: 100%

tpm %>%
  left_join(taxref, by = "orf") %>%
  replace_na(list("topphylum" = "Unidentified")) %>%
  filter(topphylum != "Unidentified") %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(sample, topphylum) %>% 
  summarise(tpm = sum(tpm), .groups = 'drop') %>%
  # Call the plot
  ggplot(aes(x = sample, 
             y = tpm, 
             fill = fct_relevel(topphylum, rev(names(cols.order)))
             )
         ) +
  labs(x = '', y = expression('Relative abundance (tpm '%*% '10'^'6'*')'), fill = "Order") +
  geom_col() +
  scale_y_continuous(breaks = c(0, 2e5, 4e5, 6e5, 8e5, 1e6),
                labels = c("0.0", "0.2", "0.4", "0.6", "0.8", "1.0")
                ) +
  scale_fill_manual(values = cols.order) +
  theme_tidy(ratio = 0.5) + 
  theme(
    axis.ticks.x = element_blank(),
    legend.key.height = unit(3.5, "mm"),
    legend.margin = margin(),
    panel.border = element_rect(fill = "transparent", linewidth = 0.5),
    axis.text.x = element_blank()
    ) +
  annotate("segment", x =  0.55, xend =  4.45, y = -25000, yend = -25000) +
  annotate("segment", x =  4.55, xend =  8.45, y = -25000, yend = -25000) +
  annotate("segment", x =  8.55, xend = 12.45, y = -25000, yend = -25000) +
  annotate("segment", x = 12.55, xend = 16.45, y = -25000, yend = -25000) +
  annotate("segment", x = 16.55, xend = 20.45, y = -25000, yend = -25000) +
  annotate("label", x = 2.5, y = -25000, label = "Station 5", label.size = 0, size = 7/.pt) +
  annotate("label", x = 6.5, y = -25000, label = "26", label.size = 0, size = 7/.pt) +
  annotate("label", x =10.5, y = -25000, label = "42", label.size = 0, size = 7/.pt) +
  annotate("label", x =14.5, y = -25000, label = "53", label.size = 0, size = 7/.pt) +
  annotate("label", x =18.5, y = -25000, label = "56", label.size = 0, size = 7/.pt)
```


```{r}
#| label: table-genus
#| tbl-cap: Twenty most abundant genera over all samples, based on tpm. The order and phylum of the respective genera are also displayed.  

t <- tpm %>%
  inner_join(tax, by = "orf") %>%
  group_by(genus, sample) %>%
  # Sum the abundance of each phylum within a sample
  summarise(tpm = sum(tpm), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean.tpm = sum(tpm), .groups = 'drop') %>%
  filter(!is.na(genus)) %>%
  top_n(20, mean.tpm)

t %>% left_join(tax, by = "genus") %>% 
  select(genus, order, phylum, mean.tpm) %>% 
  distinct() %>%
  arrange(desc(mean.tpm)) %>%
  rename(Genus = genus, Order = order, Phylum = phylum) %>%
  knitr::kable()
```


```{r, include=FALSE}
t <- tpm %>%
  inner_join(tax, by = "orf") %>%
  group_by(genus, sample) %>%
  # Sum the abundance of each phylum within a sample
  summarise(tpm = sum(tpm), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean.tpm = sum(tpm), .groups = 'drop') %>%
  filter(!is.na(genus)) %>%
  top_n(10, mean.tpm) %>% pull(genus)
```


@fig-genus shows the most abundant genera. It reveals how some genera (e.g. Pelagibacter and Ketobacter) are more abundant (and active) in the 10 m depth, while for example Nitrosopumilus is more abundant in the deeper sites. Note from the metadata that the depth of chlorophyll max strongly varies between the stations.


```{r}
#| label: fig-genus
#| fig-cap: Community composition of all RNA transcripts on genus level. Only the top ten most abundant genera are shown, accounting for ~ 20 % of the tpm on average.
#| out-width: 100%


tpm %>%
  left_join(tax, by = "orf") %>% filter(genus %in% t) %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(sample, genus) %>% 
  summarise(tpm = sum(tpm), .groups = 'drop') %>%
  # Call the plot
  ggplot(aes(x = sample, 
             y = tpm, 
             fill = fct_relevel(genus, rev(names(cols.genus)))
             )
         ) +
  labs(x = '', y = expression('Relative abundance (tpm '%*% '10'^'6'*')'), fill = "Genus") +
  geom_col() +
  scale_y_continuous(breaks = c(0, 2e5, 4e5, 6e5, 8e5, 1e6),
                labels = c("0.0", "0.2", "0.4", "0.6", "0.8", "1.0")
                ) +
  scale_fill_manual(values = cols.genus) +
  theme_tidy(ratio = 0.5) + 
  theme(
    axis.ticks.x = element_blank(),
    legend.key.height = unit(3.5, "mm"),
    legend.margin = margin(),
    panel.border = element_rect(fill = "transparent", linewidth = 0.5),
    axis.text.x = element_blank()
    ) +
  annotate("segment", x =  0.55, xend =  4.45, y = -10000, yend = -10000) +
  annotate("segment", x =  4.55, xend =  8.45, y = -10000, yend = -10000) +
  annotate("segment", x =  8.55, xend = 12.45, y = -10000, yend = -10000) +
  annotate("segment", x = 12.55, xend = 16.45, y = -10000, yend = -10000) +
  annotate("segment", x = 16.55, xend = 20.45, y = -10000, yend = -10000) +
  annotate("label", x = 2.5, y = -10000, label = "Station 5", label.size = 0, size = 7/.pt) +
  annotate("label", x = 6.5, y = -10000, label = "26", label.size = 0, size = 7/.pt) +
  annotate("label", x =10.5, y = -10000, label = "42", label.size = 0, size = 7/.pt) +
  annotate("label", x =14.5, y = -10000, label = "53", label.size = 0, size = 7/.pt) +
  annotate("label", x =18.5, y = -10000, label = "56", label.size = 0, size = 7/.pt)
```



## Beta diversity using NMDS (all stations)

As the metatranscriptomes from station 5 seem off (especially the duplicates), the ordination is run with (a) and without (b) this station. On a first glance it seems that the samples from 10 m depth and chlorophyll max. have been mixed up. The NMDS is run with default parameters.

```{r run the nmds, results='hide', message=FALSE}
set.seed(999)

nmds <- tpm %>%
  # Sum the abundance of both coverage per metagenome
  inner_join(eggnog, by = "orf") %>%
  # Filter out NA and add multiple KOs on new line
  filter(kegg_module != "-") %>% separate_rows(kegg_module, sep = ",") %>%
  # Sum the abundance of each KO within the metagenomes
  group_by(sample, kegg_module) %>% 
  summarise(tpm = sum(tpm), .groups = "drop") %>%
  # Pivot to prepare a numerical matrix
  pivot_wider(names_from = "kegg_module", values_from = "tpm", values_fill = 0) %>%
  column_to_rownames("sample") %>%
  vegan::metaMDS()

nmds
```


```{r extract nmds scores}
nmds.scores <- vegan::scores(nmds)$sites %>%
  as.data.frame() %>%
  rownames_to_column("sample")
```


```{r plot the nmds, fig.width=12/2.53}
p1 <- nmds.scores %>%
  inner_join(meta, by = "sample") %>%
  mutate(station = factor(station, levels = names(cols.station))) %>%
  # Plot
  ggplot(aes(x = NMDS1, 
             y = NMDS2, 
             colour = station,
             shape = depth
             )) +
  geom_vline(xintercept = 0, linetype = 'dotted', linewidth = 0.25) +
  geom_hline(yintercept = 0, linetype = 'dotted', linewidth = 0.25) +
  # Points for samples, coloured by site
  geom_point(size = 2) +
  scale_shape_discrete(labels = c("10 m", "Chl. max"), name = "") +
  scale_color_manual(values = cols.station, name = "Station") +
  theme_tidy() +
  annotate('text', x = -Inf, y = -Inf, size = 7/.pt,
           label = paste('Stress = ', round(nmds$stress, digits = 2)),
           hjust = -0.1, vjust = -0.5) +
  theme(
    axis.title.y = element_text(margin = margin()),
    panel.border = element_rect(linewidth = 0.75, colour = "black", fill = NA),
    legend.text = element_text(margin = margin(l = 0)),
    legend.box.margin = margin(l = -8)
    )
```


## Beta diversity using NMDS

In this NMDS, the replicates from station 5 (S1 and S2) are removed prior to running the NMDS.

```{r run the nmds without station five, results='hide', message=FALSE}
set.seed(555)

nmds <- tpm %>%
  filter(!sample %in% c("S1.1", "S1.2", "S2.1", "S2.2")) %>%
  # Sum the abundance of both coverage per metagenome
  inner_join(eggnog, by = "orf") %>%
  # Filter out NA and add multiple KOs on new line
  filter(kegg_module != "-") %>% separate_rows(kegg_module, sep = ",") %>%
  # Sum the abundance of each KO within the metagenomes
  group_by(sample, kegg_module) %>% 
  summarise(tpm = sum(tpm), .groups = "drop") %>%
  # Pivot to prepare a numerical matrix
  pivot_wider(names_from = "kegg_module", values_from = "tpm", values_fill = 0) %>%
  column_to_rownames("sample") %>%
  vegan::metaMDS()

nmds
```


```{r}
nmds.scores <- vegan::scores(nmds)$sites %>%
  as.data.frame() %>%
  rownames_to_column("sample")
```


```{r fig.width=12/2.53}
p2 <- nmds.scores %>%
  inner_join(meta, by = "sample") %>%
  # Plot
  ggplot(aes(x = NMDS1, 
             y = NMDS2, 
             colour = station,
             shape = depth
             )) +
  geom_vline(xintercept = 0, linetype = 'dotted', linewidth = 0.25) +
  geom_hline(yintercept = 0, linetype = 'dotted', linewidth = 0.25) +
  # Points for samples, coloured by site
  geom_point(size = 2) + ylab("") +
  scale_shape_discrete(labels = c("10 m", "Chl. max"), name = "", guide = "none") +
  scale_color_manual(values = cols.station, name = "Station", guide = "none") +
  theme_tidy() +
  stat_ellipse(aes(x = NMDS1, y = NMDS2, colour = depth), 
               geom = "polygon", show.legend = F, level = 0.90,
               fill = NA, linetype = "dashed", color = "black", linewidth = 0.2) +
  annotate('text', x = -Inf, y = -Inf, size = 7/.pt,
           label = paste('Stress = ', round(nmds$stress, digits = 2)),
           hjust = -0.1, vjust = -0.5) +
  theme(
    axis.title.y = element_text(margin = margin()),
    panel.border = element_rect(linewidth = 0.75, colour = "black", fill = NA),
    legend.text = element_text(margin = margin(l = 0)),
    legend.box.margin = margin(l = -8)
    )
```

```{r}
library(patchwork)
```


@fig-nmds depicts the results on the NMDS for all stations (a) and for the NMDS with station 5 removed (b). The shape of the sample depicts the depth of the sample. Note that there is a considerable variations among the depth of chlorophyll max. For example, the chl. max. of station 56 is at 20 m, while chl. max of station 42 is at 45 m depth. Possibly this could be an explanation why the four samples of the two depths from station 56 are more similar?


```{r}
#| label: fig-nmds
#| out-width: 100%
#| fig-cap: NMDS on all orthologous genes (KO). a NMDS on all five stations. b NMDS after ommitting station 5. The ellipse depicts a 90 % confidence interval. 


p1 + p2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = c("a", "b")) & theme(
  plot.margin = margin(),
  plot.tag.position = c(0.1, 1.05)
)
```


## Functional annotation - most abundant KOs


```{r}
#| tbl-cap: The most abundant KEGG orthologs (KO), their description (from Eggnog), and the summed tpm of each KO.
 
eggnog %>%
  filter(kegg_ko != "-") %>%
  inner_join(tpm, by = "orf") %>%
  group_by(kegg_ko, description) %>% summarise(tpm = sum(tpm), .groups = "drop") %>%
  slice_max(tpm, n = 20) %>%
  knitr::kable(digits = 0, format = "simple")
```



The gene list is from Royo-Llonch et al. [-@royo2021compendium]. It contains the KEGG ortholog (KO) and the pathway the gene is involved in. KOs of interest in the data that are not represented in the gene list are manually categorized. @fig-cfix shows the RNA transcripts for the carbon fixation pathways for each depth layer. To do so, one KO catalyzing the rate-limiting reaction was selected for each pathway. The Wood-Ljungdahl pathway was also included in this analysis but the representing KO was not detected, or the detected KO did not have a taxonomic annotation. 


```{r, include=FALSE}
genelist <- read_tsv("data/genelist.txt", show_col_types = FALSE) %>%
  rename_with(tolower) %>%
  rename(kegg_ko = ko_id)
```


```{r, fig.width=10/2.53}
#| fig-cap: The six carbon fixation pathways and the summed tpm from all of the samples (n = 10 for each depth layer). These pathways were categorized based on KOs with one KO per pathway as a marker gene.
#| label: fig-cfix
#| out-width: 100%

eggnog %>%
  separate_longer_delim(kegg_ko, delim = ",") %>%
  inner_join(genelist, by = "kegg_ko") %>%
  mutate(pathway = case_when(
    pathway == "Reductive acetyl-CoA pathway (Wood-Ljungdahl pathway)" ~ "Wood-Ljungdahl pathway",
    pathway == "3-Hydroxypropionate bi-cycle" ~ "3-hydroxypropionate bi-cycle",
    pathway == "Hydroxypropionate-hydroxybutylate cycle" ~ "HP/HB cycle",
    pathway == "Reductive citrate cycle (rTCA cycle; Arnon-Buchanan cycle)" ~ "Reverse TCA cycle",
    pathway == "Hydroxypropionate-hydroxybutylate cycle/ Dicarboxylate-hydroxybutyrate cycle" ~ "DC/HB cycle",
    pathway == "Calvin Cycle" ~ "Calvin cycle",
    .default = pathway
  )) %>%
  filter(kegg_ko %in% c(
    "K01601", # ribulose-bisphosphate carboxylase large chain - EC:4.1.1.39 - Calvin cycle
    "K09709", # 3-methylfumaryl-CoA hydratase - EC:4.2.1.153 - 3-hydroxypropionate bi-cycle
    "K15019", # 3-hydroxypropionyl-coenzyme A dehydratase - EC:4.2.1.116 - HP/HB cycle
    "K00175", # 2-oxoglutarate/2-oxoacid ferredoxin oxidoreductase subunit beta - EC:1.2.7.3 1.2.7.11 - Reverse TCA cycle
    "K14534", # 4-hydroxybutyryl-CoA dehydratase / vinylacetyl-CoA-Delta-isomerase - EC:4.2.1.120 5.3.3.3 - DC/HB cycle
    "K00198"  # anaerobic carbon-monoxide dehydrogenase catalytic subunit - EC:1.2.7.4 - Wood-Ljungdahl
  )) %>%
  inner_join(tax, by = "orf") %>% inner_join(tpm, by = "orf", relationship = "many-to-many") %>% inner_join(meta, by = "sample") %>%
  group_by(pathway, domain, depth) %>% summarise(tpm = sum(tpm), .groups = "drop") %>%
  mutate(depth = factor(depth, levels = c("chlMax", "10m"))) %>%
  filter(domain %in% c("Bacteria", "Archaea")) %>%
  ggplot(aes(
    x = tpm,
    y = fct_reorder(pathway, tpm),
    group = depth,
    fill = depth
  )) +
  geom_col(position = position_dodge(width = 0.95)) + labs(x = "Relative abundance (tpm)", y = "") +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_fill_manual(values = c("#046C9A", "#78B7C5"), 
                    labels = c("10m" = "10 m", "chlMax" = "Chl. max")
                    ) +
  annotate("text", x = 3500, y = 3.75, label = "98.3 % Nitrosopumilus sp.", hjust = 0, size = 7 / .pt) + # Chl. max
  annotate("text", x = 1000, y = 4.26, label = "97.8 % Nitrosopumilus sp.", hjust = 0, size = 7 / .pt) + # 10 m
  theme_tidy(fontsize = 7, ratio = 0.8) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.2),
    axis.ticks.y = element_blank(),
    legend.title = element_blank()
  )
```


¨

# Differential gene expression using edgeR

This analysis helps to identify differentially expressed genes between the 10 m depth and the chlorophyll max depth. Station 5 is not included in the analysis as it introduces too much noise.

```{r}
library(edgeR)

m <- tpm %>%
  filter(!sample %in% c("S1.1","S1.2","S2.1","S2.2")) %>% # Leave out station 5 for the moment
  select(-tpm) %>%
  arrange(sample) %>%
  pivot_wider(values_from = "count", names_from = "sample", values_fill = 0) %>%
  column_to_rownames("orf") # Create a numeric matrix

dge <- tpm %>%
  filter(!sample %in% c("S1.1","S1.2","S2.1","S2.2")) %>% # Leave out station 5 for the moment
  select(-tpm) %>%
  inner_join(eggnog, by = "orf") %>% separate_longer_delim(kegg_ko, delim = ",") %>%
  group_by(sample, kegg_ko) %>% summarise(count = sum(count), .groups = "drop") %>%
  filter(kegg_ko != "-") %>%
  arrange(sample) %>%
  pivot_wider(values_from = "count", names_from = "sample", values_fill = 0) %>%
  column_to_rownames("kegg_ko") %>%
  DGEList(group = meta %>% filter(station != 5) %>% pull(depth), lib.size = colSums(m))
```

The chunk below is to create the design matrix, make the contract, and run the actual analysis using the functions glmQLfit() and glmQLFTest()

```{r}
design.mat <- model.matrix(~ 0 + meta %>% filter(station != 5) %>% arrange(sample) %>% pull(depth))
rownames(design.mat) <- colnames(m)
colnames(design.mat) <- c("shallow", "deep")
contrast <- makeContrasts(depth = shallow-deep, levels = design.mat)

dge <- dge %>%
  calcNormFactors() %>%
  estimateDisp(design.mat) %>%
  glmQLFit(design.mat, robust = TRUE)

qlf <- glmQLFTest(dge, contrast = contrast)
```

The p-values are adjusted according to the false discovery rate, and the DE genes are joined with the functional and taxonomic annotation.

```{r}
de <- topTags(qlf, adjust.method = "fdr", n = Inf) %>%
  as.data.frame() %>%
  rownames_to_column("kegg_ko") %>% rename_with(tolower) %>%
  mutate(expression = case_when(
    fdr < 0.05 & logfc > 0 ~ "up",
    fdr < 0.05 & logfc < 0 ~ "down",
    .default = "non-de"
  )) %>%
  inner_join(eggnog, by = "kegg_ko") %>%
  inner_join(tax, by = "orf") %>%
  filter(expression != "non-de") %>% tibble()
```

As the contrast was shallow vs deep, upregulated means significantly higher in the shallow samples while downregulated means higher in the deeper samples. There are `r de %>% filter(expression == "up") %>% pull(kegg_ko) %>% unique() %>% length()` KOs up regulated and `r de %>% filter(expression == "down") %>% pull(kegg_ko) %>% unique() %>% length()` KOs down regulated. `r de %>% filter(expression == "non-de") %>% pull(kegg_ko) %>% unique() %>% length()` KOs were not differential expressed.

```{r}
cols.genus <- c(
  "Pseudothioglobus" = "#C7B19C",
  "AG-430-B22" = "#A2A475",
  "Nitrosopumilus" = "#972D15",
  "Pelagibacter" = "#D69C4E",
  "Alteromonadaceae uncl." = "#FAEFD1",
  "ASP10-02a" = "#00A08A",
  "Sneathiella" = "#D8B70A",
  "AG-337-I02" = "#1B5331",
  "AG-430-B22" = "#D5D5D3",
  "Other" = "#046C9A"
)
  
cols.order <- c( # Sorted on abundance
  "Rhizobiales_B" = "#D69C4E",
  "Pelagibacterales" = "#972D15",
  "Rhodobacterales" = "#D8B70A",
  "Rhodospirillales" = "#00A08A",
  "Pseudomonadales" = "#C7B19C",
  "Rhizobiales" = "#D5D5D3",
  "HIMB59" = "#1B5331",
  "Nitrososphaerales" = "#A2A475",
  "PS1" = "#FAEFD1",
  "Other" = "#046C9A"
)
```

The 25 KOs with the lowest (up in deeper samples) and 25 with the highest log fold-change are selected. Six of these KOs are represented in the gene list, the other 19 will be manually categorized. The number of 25 was arbitrary chosen. Many of these KOs are affiliated with the archaeon Nitrosopumilus that oxidizes ammonium and is capable of autotrophic growth using the hydroxypropionate/hydroxybutyrate cycle that is more energy efficient than other aerobic C fixation pathways, including the Calvin cycle (source: https://doi.org/10.1073/pnas.1402028111). The fact that C fixation KOs affiliated with this genus are abundantly detected strongly suggests autotrophic growth, possibly due to a low concentration of organic carbon.

```{r}
top.de <- de %>%
  mutate(pathway = case_when(
    #kegg_ko == "K02035" ~ "Extracellular solute-binding protein",
    #kegg_ko == "K00412" ~ "Component of the ubiquinol-cytochrome c reductase",
    kegg_ko == "K01638" ~ "Glycolate utilization (glcB)",
    #kegg_ko == "K03646" ~ "Cell envelope biogenesis (TolA)",
    kegg_ko == "K07112" ~ "Sulphur transport",
    kegg_ko == "K02256" ~ "Cytochrome C and quinol oxidase",
    kegg_ko == "K17227" ~ "Sulfur-oxidizing protein SoxZ",
    kegg_ko == "K00394" ~ "Dissimilatory sulfate reduction (aprA)",
    kegg_ko == "K17223" ~ "Sulfur-oxidizing protein SoxX",
    kegg_ko == "K00387" ~ "Sulfite oxidation (SUOX)",
    kegg_ko == "K03520" ~ "CO oxidation (coxL, cutL)",
    kegg_ko == "K03520" ~ "CO oxidation (coxL, cutL)",
    kegg_ko == "K03518" ~ "CO oxidation (coxS)",
    kegg_ko == "K03519" ~ "CO oxidation (coxM, cutM)",
    kegg_ko == "K00363" ~ "Dissimilatory nitrite reduction (nirD)",
    kegg_ko == "K02044" ~ "Phosphonate, periplasmic substrate-binding protein",
    kegg_ko == "K02027" ~ "Carbohydrate uptake (transported)",
    kegg_ko == "K07463" ~ "Archaea-specific exonuclease",
    kegg_ko == "K15016" ~ "Hydroxypropionate-hydroxybutylate cycle",
    kegg_ko == "K21817" ~ "Synthesis of the retinal cofactor of bacteriorhodopsin",
    kegg_ko == "K07725" ~ "CRISPR-associated protein"
    #kegg_ko == "K06937" ~ "Membrane-spanning GDGT lipids",
    #kegg_ko == "K03934" ~ "NADH dehydrogenase (ubiquinone) Fe-S protein"
  )) %>% filter(!is.na(pathway))
```


```{r, include=FALSE}
top.de <- de %>%
  mutate(pathway = case_when(
    kegg_ko == "K09152" ~ "Cell division",
    kegg_ko == "K21481" ~ "Antibiotic biosynthesis",
    kegg_ko == "K07725" ~ "CRISPR-associated protein",
    kegg_ko == "K07143" ~ "Autoantigen",
    kegg_ko == "K07569" ~ "Ribosome biogenesis",
    kegg_ko == "K07463" ~ "Archaea-specific exonuclease",
    kegg_ko == "K14728" ~ "Luciferase-like monooxygenase",
    kegg_ko == "K05831" ~ "AA biosynthesis",
    kegg_ko == "K06191" ~ "Glutaredoxin-like protein",
    kegg_ko == "K07640" ~ "Signal transduction",
    kegg_ko == "K03839" ~ "Flavodoxin",
    kegg_ko == "K07458" ~ "DNA mismatch endonuclease",
    kegg_ko == "K18594" ~ "3-hydroxypropionyl-CoA synthetase",
    kegg_ko == "K06937" ~ "Membrane-spanning GDGT lipids",
    kegg_ko == "K05827" ~ "AA biosynthesis",
    kegg_ko == "K18604" ~ "Acetyl-CoA/propionyl-CoA carboxylase",
    kegg_ko == "K02044" ~ "Phosphonate, periplasmic substrate-binding protein",
    kegg_ko == "K09771" ~ "Signaling and cellular processes",
    kegg_ko == "K07464" ~ "CRISPR-associated exonuclease",
    kegg_ko == "K17325" ~ "Glycerol transporter",
    kegg_ko == "K03477" ~ "DeoR family transcriptional regulator",
    kegg_ko == "K21817" ~ "Synthesis of the retinal cofactor of bacteriorhodopsin",
    kegg_ko == "K01473" ~ "Hydantoinase oxoprolinase",
    kegg_ko == "K06963" ~ "tRNA acetyltransferase",
    kegg_ko == "K01820" ~ "Fructose and mannose metabolism",
    kegg_ko == "K15513" ~ "Benzoate degradation",
    kegg_ko == "K17324" ~ "Glycerol transporter",
    kegg_ko == "K09539" ~ "Heat shock protein",
    kegg_ko == "K05663" ~ "ABC transporter",
    kegg_ko == "K21309" ~ "Sulfite oxidation",
    kegg_ko == "K05820" ~ "3-phenylpropionic acid transporter",
    kegg_ko == "K02494" ~ "Outer membrane lipoprotein",
    kegg_ko == "K17214" ~ "Inositol transporter",
    kegg_ko == "K07639" ~ "Signal transduction, OmpR family",
    kegg_ko == "K02027" ~ "Carbohydrate uptake (transported)",
    kegg_ko == "K14153" ~ "Thiamine metabolism",
    kegg_ko == "K16877" ~ "Furfural degradation",
    kegg_ko == "K17871" ~ "Pyridine nucleotide-disulphide oxidoreductase",
    kegg_ko == "K18622" ~ "Actin binding protein",
    kegg_ko == "K14779" ~ "DEAD box helicase family",
    # In the gene list
    kegg_ko == "K15016" ~ "Hydroxypropionate-hydroxybutylate cycle",
    kegg_ko == "K08689" ~ "Biphenyl degradation",
    kegg_ko == "K16306" ~ "Glycolysis (Embden-Meyerhof pathway)",
    kegg_ko == "K02189" ~ "Cobalamin biosynthesis",
    kegg_ko == "K03934" ~ "NADH dehydrogenase (ubiquinone) Fe-S protein"
  ))

```




```{r, fig.width=14/2.53}
#| fig-cap: Differentially expressed orthologous genes (KOs) between 10 m depth and chlorophyll max (between 20 and 45 m depth). A positive fold change indicates the KO was detected in significant higher numbers at 10 m depth, while a negative value indicates higher numbers at chlorophyll max.
#| label: fig-de

top.de %>%
  filter(!is.na(order)) %>%
  mutate(order = if_else(
    order %in% names(cols.order), true = order, false = "Other"
  )) %>%
  mutate(order = factor(order, levels = names(cols.order))) %>%
  mutate(kegg_ko = paste(kegg_ko, pathway, sep = ": ")) %>%
  ggplot(aes(x = logfc, 
             y = kegg_ko, 
             fill = order
             )
         ) +
  geom_col() + geom_vline(xintercept = 0, linewidth = unit(0.5, "mm"), color = "grey") +
  labs(x = expression("Log"["2"]*" fold change"), y = "", fill = "") +
  scale_fill_manual(values = cols.order) + theme_tidy(fontsize = 6, ratio = 1.5, legend = "right") +
  theme(
    axis.ticks.y = element_blank(),
    legend.key.height = unit(3.5, "mm"),
    legend.key.spacing.y = unit(0, "mm"),
    panel.border = element_rect(fill = "transparent", linewidth = unit(0.5, "mm"))
  )
```


# C, N, P, and S cycling

```{r n.cycling}
n <- eggnog %>% 
  #separate_longer_delim(kegg_ko, delim = ",") %>% # This is equivalent to using grepl() for selecting the KOs
  filter(kegg_ko != "-") %>%
  left_join(genelist %>% filter(metabolism == "Nitrogen metabolism"), by = "kegg_ko") %>% select(kegg_ko, orf, pathway) %>%
  mutate(pathway = case_when(
  kegg_ko %in% c("K10944","K10945","K10946") ~ "Ammonia monooxygenase (AmoABC)",
  kegg_ko == "K03320" ~ "Ammonium transporter (amt)",
  kegg_ko %in% c("K02586","K02588","K02591") ~ "Nitrogen fixation (nifDHK)",
  .default = pathway
)) %>% filter(!is.na(pathway)) %>%
  inner_join(tax, by = "orf") %>% inner_join(tpm, by = "orf", relationship = "many-to-many") 
```


```{r}
n %>%
  inner_join(meta, by = "sample") %>%
  filter(order %in% names(cols.order)) %>%
  ggplot(aes(x = tpm, y = pathway, fill = order, group = depth)) +
  geom_col(position = position_dodge(width = 0.95)) + 
  scale_fill_manual(values = cols.order) +
  labs(x = "Relative abundance (tpm)", y = "") +
  theme(
    legend.title = element_blank(),
    axis.ticks.y = element_blank()
  )
```


The genes listed below are typically catalyze the rate limiting step of the pathway. Like all other analysis, filtering is done based on the KOs and grouping of the KOs is avoided.

```{r}
c <- eggnog %>%
  filter(kegg_ko != "-") %>%
  select(kegg_ko, orf) %>% mutate(pathway = case_when(
    kegg_ko == "K01637" ~ "Glyoxylate cycle: isocitrate lyase",
    kegg_ko == "K13937" ~ "Pentose phosphate pathway: G6P dehydrogenase",
    kegg_ko == "K00850" ~ "Glycolysis: PFK",
    kegg_ko == "K01690" ~ "Entner-Doudoroff pathway: phosphogluconate dehydratase",
    kegg_ko == "K00031" ~ "Citric acid cycle: isocitrate dehydrogenase",
    kegg_ko == "K18981" ~ "Uronate pathway: uronate dehydrogenase",
    kegg_ko == "K00948" ~ "PRPP biosynthesis: ribose-phosphate pyrophosphokinase",
    .default = NA
  )) %>% filter(!is.na(pathway)) %>%
  inner_join(tax, by = "orf") %>% inner_join(tpm, by = "orf", relationship = "many-to-many")

c %>%
  inner_join(meta, by = "sample") %>%
  filter(phylum %in% names(cols.phylum)) %>%
  ggplot(aes(x = tpm, y = pathway, fill = phylum, group = depth)) +
  geom_col(position = position_dodge(width = 0.95)) + scale_fill_manual(values = cols.phylum) +
  labs(x = "Relative abundance (tpm)", y = "") +
  theme(
    legend.title = element_blank(),
    axis.ticks.y = element_blank()
  )
```

```{r}
ps <- eggnog %>%
  filter(kegg_ko != "-") %>%
  select(kegg_ko, orf) %>% mutate(pathway = case_when(
    kegg_ko == "K02703" ~ "Photosystem II psbA",
    kegg_ko == "K02712" ~ "Photosystem II psbK",
    kegg_ko == "K02707" ~ "Photosystem II psbE",
    kegg_ko == "K02697" ~ "Photosystem I psaJ",
    kegg_ko == "K02274" ~ "Cytochrome c oxidase subunit I",
    kegg_ko == "K02256" ~ "Cytochrome c oxidase subunit 1",
    .default = NA
  )) %>% filter(!is.na(pathway)) %>%
  inner_join(tax, by = "orf") %>% inner_join(tpm, by = "orf", relationship = "many-to-many")   

ps %>% 
  inner_join(meta, by = "sample") %>%
  filter(phylum %in% names(cols.phylum)) %>%
  ggplot(aes(x = tpm, y = pathway, fill = phylum, group = depth)) +
  geom_col(position = position_dodge(width = 0.95)) + scale_fill_manual(values = cols.phylum) +
  labs(x = "Relative abundance (tpm)", y = "") +
  theme(
    legend.title = element_blank(),
    axis.ticks.y = element_blank()
  )
```

# DNA sequencing: 16S amplicons

```{r}
seqtab <- read_tsv("data/amplicons/ASV_table.tsv", show_col_types = FALSE) %>%
  pivot_longer(-1, names_to = "sample", values_to = "count") %>%
  filter(count > 0) %>%
  rename(seqid = ASV_ID) %>%
  group_by(sample) %>% mutate(relab = count / sum(count)) %>% ungroup()
  
asv <- read_tsv("data/amplicons/ASV_tax_species.sbdi-gtdb.tsv", show_col_types = FALSE) %>%
  rename(seqid = ASV_ID) %>% rename_with(tolower) %>% select(-species_exact, -confidence, -sequence)
```


```{r}
t <- seqtab %>%
  inner_join(asv, by = "seqid") %>%
  group_by(phylum, sample) %>%
  # Sum the abundance of each phylum within a sample
  summarise(relab = sum(relab), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean.relab = sum(relab), .groups = 'drop') %>%
  filter(!is.na(phylum)) %>%
  top_n(9, mean.relab)

cols.asvs <- c(
  "Pseudomonadota" = "#C7B19C",
  "Bacteroidota" = "#FAEFD1",
  "Thermoproteota" = "#A2A475",
  "Actinomycetota" = "#1B5331",
  "Verrucomicrobiota" = "#D5D5D3",
  "Thermoplasmatota" = "#972D15",
  "Nitrospinota" = "#D8B70A",
  "Marinisomatota" = "#00A08A",
  "Cyanobacteriota" = "#D69C4E",
  "Other" = "#046C9A"
)
```


```{r prepare the asvs taxonomy - phylum}
taxref <- asv %>%
  left_join(t %>% transmute(phylum, topphylum = phylum), by = "phylum") %>%
  replace_na(list("topphylum" = "Other"))
```


```{r, fig.width=12/2.53}
#| out-width: 100%
#| fig-cap: Community composition based on 16S ribosomal RNA gene amplicons. Showing the nine most abundant phyla while grouping the low-abundant phyla as "Other". The phyla in the legend are ordered according to overall abundance.
#| label: fig-amplicons

seqtab %>%
  inner_join(taxref, by = "seqid") %>%
  group_by(sample, topphylum) %>% summarise(relab = sum(relab) * 100, .groups = "drop") %>%
  mutate(topphylum = factor(topphylum, levels = names(cols.asvs) %>% rev())) %>%
  ggplot(aes(
    x = sample,
    y = relab,
    fill = topphylum
  )) + geom_col() +
  labs(x = "", y = "Relative abundance (%)", fill = "Phylum") + theme_tidy(ratio = 0.7) +
  scale_fill_manual(values = cols.asvs) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    legend.key.height = unit(3, "mm"),
    legend.box.margin = margin(l = -8),
    legend.text = element_text(margin = margin(l = 2, t = 2))
  ) + guides(fill = guide_legend(reverse = TRUE)) +
  annotate("segment", x =  0.55, xend =  4.45, y = -2, yend = -2) +
  annotate("segment", x =  4.55, xend =  8.45, y = -2, yend = -2) +
  annotate("segment", x =  8.55, xend = 12.45, y = -2, yend = -2) +
  annotate("segment", x = 12.55, xend = 16.45, y = -2, yend = -2) +
  annotate("segment", x = 16.55, xend = 20.45, y = -2, yend = -2) +
  annotate("label", x = 2.5, y = -2, label = "Station 5", label.size = 0, size = 7/.pt) +
  annotate("label", x = 6.5, y = -2, label = "26", label.size = 0, size = 7/.pt) +
  annotate("label", x =10.5, y = -2, label = "42", label.size = 0, size = 7/.pt) +
  annotate("label", x =14.5, y = -2, label = "53", label.size = 0, size = 7/.pt) +
  annotate("label", x =18.5, y = -2, label = "56", label.size = 0, size = 7/.pt)
```

